---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (2nd ed.), Chapter 2: Getting Started with ggplot2"
author: "Original Code: Hadley Wickham (except Matt Dancho for ggdist and tidyquant) | Modifications: Antti Rask"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2 Getting Started with ggplot2

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 2.2 Fuel Economy Data

```{r}
mpg
```

### 2.2.1 Exercises

1.
```{r}
help(mpg)

dim(mpg)

summary(mpg)

str(mpg)

glimpse(mpg)
```

2.
```{r}
data(package = "ggplot2")
```

3.
```{r}
mpg %>% 
    # Miles per gallon
    select(cty, hwy) %>% 
    mutate(
        cty_eur = 235.21 / cty,
        hwy_eur = 235.21 / hwy
    )
```

4.
```{r}
# Without touching models
mpg %>%
  distinct(manufacturer, model) %>% 
  summarize(
    n   = n(),
    .by = manufacturer 
  ) %>%
  arrange(desc(n))
```

```{r}
# With the core models only
mpg %>%
  mutate(
    core_model = str_extract(model, "^\\S+")
  ) %>%
  distinct(manufacturer, core_model) %>%
  summarize(
    n   = n(),
    .by = manufacturer
  ) %>%
  arrange(desc(n))
```

## 2.3 Key Components

```{r}
mpg %>% 
    ggplot(aes(displ, hwy)) +
    geom_point()
```

### 2.3.1 Exercises

1.
```{r}
mpg %>% 
    ggplot(aes(cty, hwy)) +
    geom_point()
```

2.
```{r}
# Models and manufacturers
mpg %>% 
  ggplot(aes(model, manufacturer)) +
  geom_point()

# One example solution with the amount of unique models by manufacturer
mpg %>%
  mutate(
    manufacturer = manufacturer %>% as.factor(),
    core_model = str_extract(model, "^\\S+")
  ) %>%
  distinct(manufacturer, core_model) %>%
  summarize(
    n   = n(),
    .by = manufacturer
  ) %>%
  
  ggplot(aes(fct_reorder(manufacturer, n), n)) +
  geom_col() +
  coord_flip() +
  labs(
    title    = "Amount of unique models by manufacturer",
    subtitle = "",
    x        = "",
    y        = ""
  )
```

3.
```{r}
# 1.
mpg %>% 
  ggplot(aes(cty, hwy)) +
  geom_point()

# 2.
diamonds %>% 
  ggplot(aes(carat, price)) +
  geom_point()

# 3.
economics %>%
  ggplot(aes(date, unemploy)) +
  geom_line()

# 4.
mpg %>% 
  ggplot(aes(cty)) +
  geom_histogram()
```

## 2.4 Color, Size, Shape and Other Aesthetic Attributes

```{r}
mpg %>%
ggplot(aes(displ, cty, color = class)) +
    geom_point()
```

```{r}
p <- mpg %>% 
  ggplot(aes(displ, hwy))
  
p +
  geom_point(aes(color = "blue"))

p +
  geom_point(color = "blue")
```


### 2.4.1 Exercises

1.
```{r}
# Map color to continuous value
mpg %>% 
  ggplot(aes(cty, hwy, color = displ)) +
  geom_point()

# Map color to categorical value
mpg %>% 
  ggplot(aes(cty, hwy, color = trans)) +
  geom_point()

# Use more than one aesthetic in a plot
mpg %>% 
  ggplot(aes(cty, hwy, color = trans, size = trans)) +
  geom_point()
```

2.
```{r}
# Commented out, because causes error and can't see the other plot with the error:
# mpg %>%
#   ggplot(aes(cty, hwy, shape = displ)) +
#   geom_point()

mpg %>% 
  ggplot(aes(cty, hwy, shape = trans)) +
  geom_point()
```

3.
```{r}
mpg %>%
    ggplot(aes(drv, cty)) +
    geom_boxplot()

mpg %>%
    ggplot(aes(class, displ)) +
    geom_point(aes(color = drv))
```

## 2.5 Faceting

```{r}
mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_point() +
  facet_wrap(vars(class))
```

### 2.5.1. Exercises

1.
```{r}
p <- mpg %>% 
  ggplot(aes(drv, displ, fill = class)) +
  geom_col(position = "dodge")

p +
  facet_wrap(vars(hwy))

p +
  facet_wrap(vars(cyl))
```

2.
```{r}
mpg %>% 
  ggplot(aes(displ, cty)) +
  geom_point() +
  facet_wrap(vars(cyl))
```

3.
```{r}
p <- mpg %>% 
  ggplot(aes(displ, cty)) +
  geom_point()

p +
  facet_wrap(
    vars(cyl),
    nrow = 4,
    ncol = 1
  )

p +
  facet_wrap(
    vars(cyl),
    nrow = 1,
    ncol = 4
  )
```

## 2.6 Plot Geoms

### 2.6.1 Adding a Smoother to a Plot

```{r}
mpg %>% 
  ggplot(aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()
```

```{r}
# How span affects the resulting smoother
p <- mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_point()

p +
  geom_smooth(span = 0.2)

p + 
  geom_smooth(span = 1)
```

#### mgcv

```{r}
library(mgcv)

mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth(method = "gam", formula = y ~ s(x))
```

```{r}
# Line of best fit
mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

### 2.6.2 Boxplots and jittered points

```{r}
p <- mpg %>%
  ggplot(aes(drv, hwy)) 
  
p +
  geom_point()

p +
  geom_jitter()

p +
  geom_boxplot()

p +
  geom_violin()
```

### 2.6.3 Histograms and frequency polygons

```{r}
p <- mpg %>%
  ggplot(aes(hwy))

p +
  geom_histogram()

p +
  geom_freqpoly()

p + 
  geom_freqpoly(binwidth = 2.5)

p + 
  geom_freqpoly(binwidth = 1)
```

```{r}
mpg %>% 
  ggplot(aes(displ, color = drv)) + 
  geom_freqpoly(binwidth = 0.5)

mpg %>% 
  ggplot(aes(displ, fill = drv)) + 
  geom_histogram(binwidth = 0.5) + 
  facet_wrap(vars(drv), ncol = 1)
```

### 2.6.4 Bar charts

```{r}
mpg %>% 
  ggplot(aes(manufacturer)) + 
  geom_bar()
```

```{r}
drugs <- tibble(
  drug = c("a", "b", "c"),
  effect = c(4.2, 9.7, 6.1)
)

p <- drugs %>% 
  ggplot(aes(drug, effect))
  
p +
  geom_bar(stat = "identity")

p +
  geom_point()
```

### 2.6.5 Time series with line and path plots

```{r}
economics %>%
  ggplot(aes(date, unemploy / pop)) +
  geom_line()

economics %>%
  ggplot(aes(date, uempmed)) +
  geom_line()
```

#### lubridate

```{r}
library(lubridate)

# I decided to replace all of this... 
# year <- function(x) as.POSIXlt(x)$year + 1900
# ... with the year() function from lubridate

p <- economics %>% 
  ggplot(aes(unemploy / pop, uempmed))

p +
  geom_path() +
  geom_point()

p +
  geom_path(color = "grey50") +
  geom_point(aes(color = year(date)))
```

### 2.6.6 Exercises

1.
```{r}
mpg %>%
  ggplot(aes(cty, hwy)) +
  geom_jitter(alpha = 0.5)
```

2.
```{r}
mpg %>%
  ggplot(aes(class, hwy)) +
  geom_boxplot()

mpg %>%
  ggplot(aes(reorder(class, hwy), hwy)) +
  geom_boxplot()
```

3.
```{r}
diamonds %>% 
  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.3)
```

4.
```{r}
diamonds %>% 
  ggplot(aes(fct_reorder(cut, price), price)) +
  geom_boxplot()

diamonds %>% 
  ggplot(aes(price, after_stat(density), color = cut)) +
  geom_freqpoly(binwidth = 200)
```

6.
```{r}
?geom_bar()

p <- mpg %>%
  ggplot(aes(class))

p +
  geom_bar()

p +
  geom_bar(aes(weight = displ))
```

## 2.7 Modifying the axes

```{r}
p <- mpg %>%
  ggplot(aes(cty, hwy)) +
  geom_point(alpha = 1 / 3)

p

p +
  labs(
    x = "city driving (mpg)",
    y = "highway driving (mpg)"
  )

# Remove the axis labels with NULL
p +
  labs(
    x = NULL,
    y = NULL
  )
```

```{r}
p <- mpg %>%
  ggplot(aes(drv, hwy)) +
  geom_jitter(width = 0.25)

p

p + 
  xlim("f", "r") + 
  ylim(20, 30)

# For continuous scales, use NA to set only one limit
mpg %>%
  ggplot(aes(drv, hwy)) +
  geom_jitter(width = 0.25, na.rm = TRUE) + 
  ylim(NA, 30)
```

## 2.8 Output

```{r}
p <- mpg %>%
  ggplot(aes(displ, hwy, color = factor(cyl))) +
  geom_point()
```

```{r}
p

print(p)
```

```{r}
# Save png to disk
ggsave("img/plot.png", p, width = 5, height = 5)
```

```{r}
summary(p)
```

```{r}
saveRDS(p, "output/plot.rds")
q <- readRDS("output/plot.rds")
q
```

## 2.X Raincloud plot with ggdist

https://www.business-science.io/r/2021/07/22/ggdist-raincloud-plots.html

```{r}
library(ggdist)
library(tidyquant)

mpg %>%
  ggplot(aes(factor(drv), hwy, fill = factor(drv))) +
  
  # Add half-violin from {ggdist} package
  stat_halfeye(
    # Custom bandwidth
    adjust        = 0.5,
    # Move geom to the right
    justification = -.2,
    # Remove slab interval
    .width        = 0,
    point_color   = NA
  ) +
    
  # Add boxplot
  geom_boxplot(
    width          = .12,
    # Remove outliers
    outlier.color  = NA,
    alpha          = 0.5
  ) +
  
  # Add dot plots from {ggdist} package
  stat_dots(
    # Orientation to the left
    side = "left",
    # Move geom to the left
    justification = 1.1,
    # Adjust grouping (binning) of observations
    binwidth = .25
  ) +
    
  # Adjust theme
  scale_fill_tq() +
  theme_tq() +
  labs(
    title    = "Raincloud plot",
    subtitle = "Is it better to get all the information in one plot?",
    x        = "Drivetrain Type",
    y        = "Highway Fuel Economy (MPG)",
    fill     = "Drivetrain"
  ) +
  coord_flip()
```
