---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (2nd ed.), Chapter 7 - Positioning"
author: "Original Code: Hadley Wickham | Modifications: Antti Rask"
date: "2023-01-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 7 Positioning

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 7.2 Facetting

```{r}
mpg2 <- mpg %>%
  # subset -> filter
  filter(cyl != 5 & drv %in% c("4", "f") & class != "2seater")
```

### 7.2.1 Facet Wrap

```{r}
base <- mpg2 %>% 
  ggplot(aes(displ, hwy)) + 
  geom_blank() + 
  labs(
    x = NULL,
    y = NULL
  )

base +
  facet_wrap(
    vars(class),
    ncol = 3
  )

base + facet_wrap(
  vars(class),
  ncol     = 3,
  as.table = FALSE
)

base +
  facet_wrap(
    vars(class),
    nrow = 3
  )

base +
  facet_wrap(
    vars(class),
    nrow = 3,
    dir  = "v"
  )
```

### 7.2.2 Facet Grid

```{r}
base +
  facet_grid(cols = vars(cyl))

base +
  facet_grid(rows = vars(drv))

base +
  facet_grid(vars(drv), vars(cyl))
```

### 7.2.3 Controlling Scales

```{r}
p <- mpg2 %>% 
  ggplot(aes(cty, hwy)) +
  geom_abline() +
  geom_jitter(width = 0.1, height = 0.1) 

p + 
  facet_wrap(vars(cyl))

p +
  facet_wrap(vars(cyl), scales = "free")
```

```{r}
economics_long

economics_long %>%
  ggplot(aes(date, value)) + 
  geom_line() + 
  facet_wrap(
    vars(variable),
    scales = "free_y",
    ncol   = 1
  )
```

```{r}
mpg3 <- mpg2 %>%
  mutate(
    model        = as_factor(model),
    manufacturer = as_factor(manufacturer)
  )

mpg3 %>%
  ggplot(aes(cty, fct_reorder(model, cty))) + 
  geom_point() +
  facet_grid(
    rows = vars(
      fct_reorder(
        manufacturer,
        cty,
        .desc = TRUE
      )
    ),
    scales = "free",
    space  = "free"
  ) +
  theme(strip.text.y = element_text(angle = 0)) +
  labs(y = "model")
```

### 7.2.4 Missing Facetting Variables

```{r}
tbl1 <- tibble(
  x      = 1:3,
  y      = 1:3,
  gender = c("f", "f", "m")
)

tbl2 <- tibble(x = 2, y = 2)

tbl1 %>%
  ggplot(aes(x, y)) + 
  geom_point(
    data  = tbl2,
    color = "red",
    size  = 2
  ) + 
  geom_point() + 
  facet_wrap(vars(gender))
```

### 7.2.5 Grouping vs. Facetting

```{r}
tbl <- data.frame(
  x = rnorm(120, c(0, 2, 4)),
  y = rnorm(120, c(1, 2, 1)),
  z = letters[1:3]
) %>% as_tibble()

# Grouping
p <- tbl %>%
  ggplot(aes(x, y))

p +
  geom_point(aes(color = z)) +
  labs(title = "Grouping")

# Faceting
p +
  geom_point() + 
  facet_wrap(vars(z)) +
  labs(title = "Faceting")

# Faceting with means
tbl_sum <- tbl %>%
  summarize(
    x   = mean(x),
    y   = mean(y),
    .by = z
  ) %>%
  rename(z2 = z)

p + 
  geom_point() + 
  geom_point(
    data       = tbl_sum,
    aes(color  = z2),
    size       = 4
  ) + 
  facet_wrap(vars(z)) +
  labs(title = "Faceting with Means")

# Grouping + faceting
tbl3 <- tbl %>% 
  select(-z)

p +
  geom_point(data = tbl3, color = "grey70") +
  geom_point(aes(color = z)) + 
  facet_wrap(vars(z)) +
  labs(title = "Grouping + Faceting")
```

### 7.2.6 Continuous Variables

```{r}
mpg4 <- mpg2 %>% 
  mutate(
    # Bins of width 1
    disp_w = displ %>% cut_width(1),
    # Six bins of equal length
    disp_i = displ %>% cut_interval(6),
    # Six bins containing equal numbers of points
    disp_n = displ %>% cut_number(6)
  )

plot <- mpg4 %>% 
  ggplot(aes(cty, hwy)) +
  geom_point() +
  labs(
    x = NULL,
    y = NULL
  )

plot +
  facet_wrap(vars(disp_w), nrow = 1)

plot +
  facet_wrap(vars(disp_i), nrow = 1)

plot +
  facet_wrap(vars(disp_n), nrow = 1)
```

### 7.2.7 Exercises

1.
```{r}
carats <- diamonds %>%
  mutate(carat = carat %>% as.integer())

carats %>% 
  ggplot(
    aes(
      price,
      color = as.factor(carat)
    )
  ) +
  geom_freqpoly() +
  facet_wrap(vars(cut), scales = "free_y") +
  labs(color = "carat")

carats %>%
  ggplot(aes(price, color = cut)) +
  geom_freqpoly() +
  facet_wrap(vars(carat), scales = "free_y")
```

2.
```{r}
p <- diamonds %>%
  ggplot(aes(carat, price, color = color)) +
  geom_point()

p

p +
  facet_wrap(vars(color))

diamonds2 <- diamonds %>% 
  select(-color)

diamonds %>%
  ggplot(aes(carat, price)) +
  geom_point(data = diamonds2, color = "grey70") +
  geom_point(aes(color = color)) + 
  facet_wrap(vars(color))
```

4.
```{r}
mpg2 %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  geom_smooth(
    data = mpg2 %>% select(-class),
    se = FALSE
  ) +
  facet_wrap(vars(class))
```

## 7.4 Linear Coordinate Systems

### 7.4.1 Zooming into a Plot with coord_cartesian()

```{r}
base <- mpg %>% 
  ggplot(aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()

# Full dataset
base

# Scaling to 4--6 throws away data outside that range
base +
  scale_x_continuous(limits = c(4, 6))

# Zooming to 4--6 keeps all the data but only shows some of it
base +
  coord_cartesian(xlim = c(4, 6))
```

### 7.4.2 Flipping the Axes with coord_flip()

```{r}
p <- mpg %>%
  ggplot(aes(displ, cty)) +
  geom_point() + 
  geom_smooth()

p

# Exchanging cty and displ rotates the plot 90 degrees, # but the smooth is fit to the rotated data.
mpg %>%
  ggplot(aes(cty, displ)) + 
  geom_point() + 
  geom_smooth()

# coord_flip() fits the smooth to the original data, and then rotates the output
p +
  coord_flip()
```

## 7.5 Non-linear Coordinate Systems

```{r}
rect <- tibble(x = 50, y = 50)
line <- tibble(x = c(1, 200), y = c(100, 1))

base <- ggplot(mapping = aes(x, y)) +
  geom_tile(
    data = rect,
    aes(width = 50, height = 50)
  ) + 
  geom_line(data = line) + 
  labs(
    x = NULL,
    y = NULL
  )

base +
  labs(title = "base")

base +
  coord_polar("x") +
  labs(title = 'coord_polar("x")')

base +
  coord_polar("y") +
  labs(title = 'coord_polar("y")')

base +
  coord_flip() +
  labs(title = "coord_flip")

base +
  coord_trans(y = "log10") +
  labs(title = 'coord_trans(y = "log10")')

base +
  coord_fixed() +
  labs(title = "coord_fixed()")
```

Munching:

1. We start with a line parameterised by its two endpoints:
```{r}
tbl <- tibble(
  r     = c(0, 1),
  theta = c(0, 3 / 2 * pi)
)

tbl %>%
  ggplot(aes(r, theta)) + 
  geom_line() + 
  geom_point(size = 2, color = "red")
```

2. We break it into multiple line segments, each with two endpoints:
```{r}
interp <- function(rng, n) {
  seq(rng[1], rng[2], length = n)
}

munched <- tibble(
  r     = interp(tbl$r, 15),
  theta = interp(tbl$theta, 15)
)

munched %>%
  ggplot(aes(r, theta)) + 
  geom_line() + 
  geom_point(size = 2, color = "red")
```

3. We transform the locations of each piece:
```{r}
transformed <- transform(
  munched,
  x = r * sin(theta),
  y = r * cos(theta)
)

transformed %>%
  ggplot(aes(x, y)) + 
  geom_path() + 
  geom_point(size = 2, color = "red") + 
  coord_fixed()
```

### 7.5.1 Transformations with coord_trans()

```{r, warning=FALSE}
library(scales)

# Linear model on original scale is poor fit
base <- diamonds %>% 
  ggplot(aes(carat, price)) + 
  stat_bin2d() + 
  geom_smooth(method = "lm") + 
  labs(
    x = NULL,
    y = NULL
  ) +
  theme(legend.position = "none")

base

# Better fit on log scale, but harder to interpret
base +
  scale_x_log10() + 
  scale_y_log10()

# Fit on log scale, then backtransform to original.
# Highlights lack of expensive diamonds with large
# carats
pow10 <- exp_trans(10)

base +
  scale_x_log10() + 
  scale_y_log10() + 
  coord_trans(x = pow10, y = pow10)
```

### 7.5.2 Polar coordinates with coord_polar()

```{r}
base <- mtcars %>% 
  ggplot(aes(factor(1), fill = factor(cyl))) +
  geom_bar(width = 1) +
  theme(legend.position = "none") +
  scale_x_discrete(NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, expand = c(0, 0))

# Stacked barchart
base

# Pie chart
base +
  coord_polar(theta = "y")

# The bullseye chart
base +
  coord_polar()
```

### 7.5.3 Map Projections with coord_map()

```{r}
# Prepare a map of NZ
nzmap <- map_data("nz") %>%
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  labs(
    x = NULL,
    y = NULL
  )

# Plot it in cartesian coordinates
nzmap

# With the aspect ratio approximation
nzmap +
  # coord_quickmap -> coord_sf
  coord_sf()
```

```{r}
# Since coord_map() has been superseded by coord_sf, I've modified this code to use that instead
world   <- map_data("world")
finland <- map_data("world", "finland")

worldmap <- world %>% 
  ggplot(aes(long, lat, group = group))

worldmap +
  geom_path() +
  geom_path(data = finland, color = "blue") +
  # coord_map -> coord_sf
  coord_sf()

worldmap +
  geom_polygon(fill = "white", color = "black") +
  geom_polygon(data = finland, fill = "blue") +
  # coord_map -> coord_sf
  coord_sf()
```
