---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (2nd ed.), Chapter 9 - Data Analysis"
author: "Original Code: Hadley Wickham | Modifications: Antti Rask"
date: "2023-01-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 9 Data Analysis

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 9.2 Tidy Data

Since there was no ec2 dataset ready (as it says in the book, it was derived from the economics dataset, the book just didn't show how), I thought we should start by creating it, using the economics dataset as a starting point.

```{r}
library(lubridate)

ec2 <- economics %>%
  filter(date >= "2006-01-01") %>% 
  separate(
    date,
    c("year", "month"),
    sep   = "-",
    extra = "drop"
  ) %>%
  select(year, month, uempmed) %>%
  pivot_wider(
    names_from  = year,
    values_from = uempmed
  ) %>% 
  mutate(month = month %>% as.numeric())

ec2
```

## 9.3 pivot_longer() and pivot_wider()

This was actually called Spread and Gather in the 2nd edition. However, since they have been replaced by pivot_longer and pivot_wider, I thought it would make sense to rename these sub chapters.

If you want to read more about this, you can check out these two:
pivot_longer(): https://tidyr.tidyverse.org/reference/pivot_longer.html
pivot_wider(): https://tidyr.tidyverse.org/reference/pivot_wider.html

### 9.3.1 pivot_longer()

```{r}
ec2 %>%
  pivot_longer(
    cols      = `2006`:`2015`,
    names_to  = "year",
    values_to = "unemp"
  )

ec2 %>%
  pivot_longer(
    cols      = -month,
    names_to  = "year",
    values_to = "unemp"
  )

economics_2 <- ec2 %>%
  pivot_longer(
    cols            = -month,
    names_to        = "year",
    values_to       = "rate",
    names_transform = list(year = as.integer),
    values_drop_na  = TRUE
  )

economics_2
```

```{r}
economics_2 %>% 
  ggplot(aes(year + (month - 1) / 12, rate)) +
  geom_line()

economics_2 %>%
  ggplot(aes(month, rate, group = year, color = year)) +
  geom_line(linewidth = 1)
```

### 9.3.2 pivot_wider()

```{r}
weather <- tibble(
  day = rep(1:3, 2),
  obs = rep(c("temp", "rain"), each = 3),
  val = c(c(23, 22, 20), c(0, 0, 5))
)

weather
```

```{r}
weather %>% 
  pivot_wider(
    names_from  = obs,
    values_from = val
  )
```

## 9.4 Separate and Unite

```{r}
trt <- tibble(
  # There's perhaps a temptation to switch paste0 to str_c. In this case   that would be a mistake 
  var = paste0(
    rep(c("beg", "end"), each = 3), "_", rep(c("a", "b", "c"))
  ),
  val = c(1, 4, 2, 10, 5, 11)
)

trt
```

```{r}
trt %>% 
  separate(
    var,
    c("time", "treatment"),
    "_"
  )
```

## 9.5 Case Studies

### 9.5.1 Blood Pressure

```{r}
bpd <- read_table(
  "name age start week1 week2 week3
    Anne 35 2014-03-27 100/80 100/75 120/90
    Ben 41 2014-03-09 110/65 100/65 135/70
    Carl 33 2014-04-02 125/80 <NA> <NA>",
  na = "<NA>"
)

bpd
```

```{r}
bpd_1 <- bpd %>% 
  pivot_longer(
    starts_with("week"),
    names_to  = "week",
    values_to = "bp"
  )

bpd_1
```

```{r}
bpd_2 <- bpd_1 %>%
  separate(bp, c("sys", "dia"), "/")

bpd_2
```

```{r}
bpd_3 <- bpd_2 %>% 
  extract(
    week,
    "week",
    "(\\d)",
    convert = TRUE
  ) %>% 
  arrange(name, week)

bpd_3
```

### 9.5.2 Test Scores

```{r}
scores <- tibble(
  person = rep(c("Greg", "Sally", "Sue"), each = 2),
  time   = rep(c("pre", "post"), 3),
  test1  = round(rnorm(6, mean = 80, sd = 4), 0),
  test2  = round(jitter(test1, 15), 0)
)

scores
```

```{r}
scores_1 <- scores %>% 
  pivot_longer(
    cols      = starts_with("test"),
    names_to  = "test",
    values_to = "score"
  )

scores_1
```

```{r}
scores_2 <- scores_1 %>%
  pivot_wider(
    names_from  = time,
    values_from = score
  )

scores_2
```

```{r}
scores_3 <- scores_2 %>% 
  mutate(diff = post - pre)

scores_3 %>% 
  ggplot(aes(person, diff, color = test)) +
  geom_hline(
    linewidth  = 2,
    color      = "white",
    yintercept = 0
  ) +
  geom_point() +
  geom_path(
    aes(group = person),
    color     = "grey50",
    arrow     = arrow(length = unit(0.25, "cm"))
  )
```
