---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (2nd ed.), Chapter 3 - Toolbox"
author: "Original Code: Hadley Wickham (except Taro Mieno for Raster Images, Duncan Murdoch for rgl and Dr. Dominic Royé for OpenStreetMaps) | Modifications: Antti Rask"
date: "2023-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3 Toolbox

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 3.2 Basic Plot Types

```{r}
tbl <- tibble(
  x     = c(3, 1, 5), 
  y     = c(2, 4, 6), 
  label = c("a","b","c")
)

p <- tbl %>% 
  ggplot(aes(x, y, label = label)) + 
  # Hide axis label
  labs(
    x = NULL,
    y = NULL
  ) +
  # Shrink plot title
  theme(plot.title = element_text(size = 12))

p +
  geom_point() +
  ggtitle("point")

p +
  geom_text() +
  ggtitle("text")

p +
  geom_bar(stat = "identity") +
  ggtitle("bar")

p +
  geom_tile() +
  ggtitle("raster")
```

```{r}
p +
  geom_line() +
  ggtitle("line")

p +
  geom_area() +
  ggtitle("area")

p +
  geom_path() +
  ggtitle("path")

p +
  geom_polygon() +
  ggtitle("polygon")
```

## 3.3 Labels

```{r}
tbl <- tibble(
  x      = 1,
  y      = 3:1,
  family = c("sans", "serif", "mono")
)

tbl %>%
  ggplot(aes(x, y)) + 
  geom_text(aes(label = family, family = family))
```

```{r}
tbl <- tibble(
  x    = 1,
  y    = 3:1,
  face = c("plain", "bold", "italic")
)

tbl %>%
  ggplot(aes(x, y)) + 
  geom_text(aes(label = face, fontface = face))
```

```{r}
tbl <- tibble(
  x    = c(1, 1, 2, 2, 1.5),
  y    = c(1, 2, 1, 2, 1.5),
  text = c(
    "bottom-left", "top-left",  
    "bottom-right", "top-right", "center"
  )
)

tbl %>%
  ggplot(aes(x, y)) +
  geom_text(aes(label = text))

tbl %>%
  ggplot(aes(x, y)) +
  geom_text(
    aes(label = text),
    vjust = "inward",
    hjust = "inward"
  )
```

```{r}
tbl <- tibble(
  trt  = c("a", "b", "c"),
  resp = c(1.2, 3.4, 2.5)
  )

tbl %>%
  ggplot(aes(resp, trt)) + 
  geom_point() + 
  geom_text(
    # paste0 -> str_c
    aes(label = str_c("(", resp, ")")),
    nudge_y   = -0.25
  ) +
  xlim(1, 3.6)
```

```{r}
mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_text(aes(label = model)) + 
  xlim(1, 8)

mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_text(aes(label = model), check_overlap = TRUE) +
  xlim(1, 8)
```

```{r}
tbl <- tibble(
  waiting   = c(55, 80), 
  eruptions = c(2, 4.3), 
  label     = c("peak one", "peak two")
)

faithfuld %>%
  ggplot(aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) + 
  geom_label(data = tbl, aes(label = label))
```

### directlabels

```{r}
library(directlabels)

p <-
  mpg %>%
  ggplot(aes(displ, hwy, color = class))

p +
  geom_point()

p + 
  geom_point(show.legend = FALSE) +
  geom_dl(
    aes(label = class),
    method    = "smart.grid"
  )
```

## 3.4 Annotations

```{r}
economics %>% 
  ggplot(aes(date, unemploy)) + 
  geom_line()
```

```{r}
presidential_filtered <- presidential %>%
  # subset + economics$date[1] -> filter + min(economics$date)
  filter(
    start > min(economics$date),
    # added the second filter, because the economics data ends way before the presidential one
    start < max(economics$date)
  )

economics %>%
  ggplot() +
  geom_rect(
    aes(xmin = start, xmax = end, fill = party),
    ymin  = -Inf,
    ymax  = Inf,
    alpha = 0.2,
    data  = presidential_filtered
  ) +
  geom_vline(
    aes(xintercept = as.numeric(start)),
    data  = presidential_filtered,
    color = "grey50",
    alpha = 0.5
  ) +
  geom_text(
    aes(x = start, y = 2500, label = name),
    data    = presidential_filtered,
    size    = 3,
    vjust   = 0,
    hjust   = 0,
    nudge_x = 50
  ) +
  geom_line(aes(date, unemploy)) +
  scale_fill_manual(values = c("blue", "red")) +
  labs(
    x = "date",
    y = "unemployment"
  )
```

```{r}
yrng <- range(economics$unemploy)
xrng <- range(economics$date)

# paste -> str_c
caption <- str_c(
  # strwrap -> str_wrap
  str_wrap(
    "Unemployment rates in the US have varied a lot over the years",
    40
  ),
  collapse = "\n"
)

economics %>%
  ggplot(aes(date, unemploy)) + 
  geom_line() + 
  geom_text(
    aes(x, y, label = caption), 
    data = tibble(
      # xrng[1] -> min(xrng)
      x       = min(xrng),
      # yrng[2] -> max(yrng)
      y       = max(yrng),
      caption = caption
    ), 
    hjust = 0,
    vjust = 1,
    size  = 4
  )

# Same, but easier thanks to annotate()
economics %>%
ggplot(aes(date, unemploy)) + 
  geom_line() + 
  annotate(
    geom  = "text",
    # xrng[1] -> min(xrng)
    x     = min(xrng),
    # yrng[2] -> max(yrng)
    y     = max(yrng),
    label = caption,
    hjust = 0,
    vjust = 1,
    size  = 4
  )
```

```{r}
diamonds %>%
  ggplot(aes(log10(carat), log10(price))) + 
  geom_bin2d() + 
  facet_wrap(
    vars(cut),
    nrow = 1
  )
```

```{r}
mod_coef <- coef(lm(log10(price) ~ log10(carat), data = diamonds))

diamonds %>%
  ggplot(aes(log10(carat), log10(price))) + 
  geom_bin2d() +
  geom_abline(
    intercept = mod_coef[1],
    slope     = mod_coef[2],
    color     = "white",
    linewidth = 1
  ) + 
  facet_wrap(
    vars(cut),
    nrow = 1
  )
```

## 3.5 Collective Geoms

```{r}
library(nlme)

oxboys <- Oxboys

head(oxboys)
```

### 3.5.1 Multiple Groups, One Aesthetic

```{r}
oxboys %>%
  ggplot(aes(age, height, group = Subject)) + 
  geom_point() + 
  geom_line()
```

```{r}
# Don't do this!
oxboys %>%
ggplot(aes(age, height)) + 
  geom_point() + 
  geom_line()
```

### 3.5.2 Different Groups on Different Layers

```{r}
# Don't do this!
oxboys %>% 
  ggplot(aes(age, height, group = Subject)) + 
  geom_line() + 
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
oxboys %>%
  ggplot(aes(age, height)) + 
  geom_line(aes(group = Subject)) + 
  geom_smooth(
    method    = "lm",
    linewidth = 2,
    se        = FALSE
  )
```

### 3.5.3 Overriding the Default Grouping

```{r}
p <- oxboys %>%
  ggplot(aes(Occasion, height)) + 
  geom_boxplot()

p

# Don't do this!
p +
  geom_line(color = "#3366FF", alpha = 0.5)

# This works
p +
  geom_line(
    aes(group = Subject),
    color     = "#3366FF",
    alpha     = 0.5
  )
```

### 3.5.4 Matching Aesthetics to Graphic Objects

```{r}
tbl <- tibble(x = 1:3, y = 1:3, color = c(1,3,5))

# 1
tbl %>%
  ggplot(aes(x, y, color = factor(color))) + 
  geom_line(aes(group = 1), linewidth = 2) +
  geom_point(size = 5)

# 2
tbl %>%
  ggplot(aes(x, y, color = color)) + 
  geom_line(aes(group = 1), linewidth = 2) +
  geom_point(size = 5)

# 3
xgrid <- with(
  tbl,
  seq(min(x), max(x), length = 50)
)

interpolation <- tibble(
  x = xgrid,
  y = approx(
    x    = tbl$x,
    y    = tbl$y,
    xout = xgrid)$y,
  color = approx(
    tbl$x,
    tbl$color,
    xout = xgrid)$y
)

interpolation %>%
  ggplot(aes(x, y, color = color)) + 
  geom_line(linewidth = 2) +
  geom_point(data = tbl, size = 5)
```

```{r}
mpg %>%
  ggplot(aes(class)) + 
  geom_bar()

mpg %>% 
  ggplot(aes(class, fill = drv)) + 
  geom_bar()
```

```{r}
# Don't do this!
mpg %>% 
  ggplot(aes(class, fill = hwy)) + 
  geom_bar()

mpg %>%
  ggplot(aes(class, fill = hwy, group = hwy)) + 
  geom_bar()
```

### 3.5.5 Exercises

1.
```{r}
mpg %>% 
  ggplot(aes(cyl, hwy, group = cyl)) +
  geom_boxplot()
```

2.
```{r}
mpg %>%
  ggplot(aes(displ, cty, group = as.integer(displ))) + 
  geom_boxplot()
```

3.
```{r}
tbl <- tibble(x = 1:3, y = 1:3, color = c(1,3,5))

# Group = 1
p <- tbl %>%
  ggplot(aes(x, y, color = factor(color)))
  
p +
  geom_line(aes(group = 1), linewidth = 2) +
  geom_point(size = 5) +
  labs(title = "Group = 1")

# Group = 1 omitted
p + 
  geom_line(linewidth = 2) +
  geom_point(size = 5) +
  labs(title = "Group = 1 omitted")

# Group = 2
p + 
  geom_line(aes(group = 2), linewidth = 2) +
  geom_point(size = 5) +
  labs(title = "Group = 2")
```

4.
```{r}
mpg %>%
  ggplot(aes(drv)) + 
  geom_bar(color = "white")

mpg %>%
  ggplot(aes(drv, fill = hwy, group = hwy)) + 
  geom_bar(color = "white")

mpg2 <- mpg %>%
  arrange(hwy) %>%
  mutate(id = seq_along(hwy))

mpg2 %>% 
  ggplot(aes(drv, fill = hwy, group = id)) + 
  geom_bar(color = "white")
```

5.
```{r}
library(babynames)

hadley <- babynames %>% 
  filter(name == "Hadley")

hadley %>%
  ggplot(aes(year, n, group = sex, color = sex)) + 
  geom_line()
```

## 3.6 Surface Plots

```{r}
p <- faithfuld %>%
  ggplot(aes(eruptions, waiting))

p +
  # ..level.. -> after_stat(level)
  geom_contour(aes(z = density, color = after_stat(level)))

p +
  geom_raster(aes(fill = density))
```

```{r}
# Bubble plots work better with fewer observations
small <- faithfuld %>%
  slice(seq(1, nrow(.), 10))
small

# Or alternatively
small2 <- faithfuld %>%
  slice(which(row_number() %% 10 == 1))
small2

# Or alternatively
small3 <- faithfuld %>% 
   filter(row_number() %% 10 == 1)
small3

small %>%
  ggplot(aes(eruptions, waiting)) +
  geom_point(aes(size = density), alpha = 1/3) +
  scale_size_area()
```

## 3.7 Drawing Maps

### 3.7.1 Vector Boundaries

```{r}
mi_counties <- map_data("county", "michigan") %>%
  as_tibble() %>%
  select(lon = long, lat, group, id = subregion)
  
head(mi_counties)
```

```{r}
p <- mi_counties %>%
  ggplot(aes(lon, lat, group = group))
  
p +
  geom_polygon() +
  coord_sf()

p +
  geom_polygon(
    fill = NA, 
    color = "grey50"
  ) + 
  coord_sf()
```

```{r}
library(lubridate)
library(USAboundaries)

c18    <- usboundaries(date("1820-01-01"))
c18tbl <- fortify(c18) %>% as_tibble()

head(c18tbl)

c18tbl %>%
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(color = "grey50", fill = NA) +
  coord_sf()
```

### 3.7.2 Point Metadata

```{r}
library(maps)

mi_cities <- us.cities %>% 
  # tbl_df -> as_tibble
  as_tibble() %>%
  filter(country.etc == "MI") %>% 
  select(-country.etc, lon = long) %>% 
  arrange(desc(pop))

mi_cities
```

```{r}
# Don't do this!
mi_cities %>% 
  ggplot(aes(lon, lat, size = pop)) +
  geom_point() +
  scale_size_area() +
  coord_sf()
```

```{r}
mi_cities %>% 
  ggplot(aes(lon, lat)) +
  geom_polygon(
    aes(group = group),
    mi_counties,
    fill  = NA,
    color = "grey50"
  ) +
  geom_point(aes(size = pop), color = "red") +
  scale_size_area() +
  coord_sf()
```   

### 3.7.3 Raster Images

Both the old and new examples had problems (which I opened an GitHub issue about)...

So, let's try one this instead: https://tmieno2.github.io/R-as-GIS-for-Economists/create-maps.html

#### 3.7.3.1 Creating maps from sf objects

```{r}
# Read in the KS wells data
gw_KS_sf <- readRDS("input/gw_KS_sf.rds")
gw_KS_sf
```

```{r}
# Polygons
library(sf)
library(tigris)

KS_county <- counties(
  state = "Kansas",
  cb    = TRUE
) %>%
  st_as_sf() %>%
  st_transform(st_crs(gw_KS_sf))

KS_county
```

```{r}
# Points
g_wells <- gw_KS_sf %>% 
  ggplot() +
  geom_sf()

g_wells
```

```{r}
g_county <- KS_county %>%
  ggplot() +
  geom_sf()

g_county
```

```{r}
# Points
gw_Stevens <- KS_county %>%
  filter(NAME == "Stevens") %>%
  st_crop(gw_KS_sf, .)

p <- 
  gw_Stevens %>% 
  ggplot()

p +
  geom_sf(aes(color = af_used), size = 2)

p +
  geom_sf(
    aes(size = af_used),
    color    = "blue",
    shape    = 15
  )

gw_Stevens %>%
  mutate(coordinates = st_coordinates(.)) %>% 
  rowwise() %>%
  mutate(
    X           = coordinates[1],
    east_west   = case_when(
      X < -101.3 ~ "west",
      TRUE       ~ "east"
    )
  ) %>%
  ggplot() +
  geom_sf(aes(shape = east_west, color = east_west))
```

```{r}
# Polygons
KS_county %>%
  ggplot() +
  geom_sf(color = "red", fill = "darkgreen")

KS_county_with_pumping <- gw_KS_sf %>%
  filter(year == 2010) %>%
  aggregate(., KS_county, sum, na.rm = TRUE)

KS_county_with_pumping %>%
  ggplot() +
  geom_sf(aes(fill = af_used))
```

```{r}
# Plotting multiple spatial objects in one figure
ggplot() +
  geom_sf(data = gw_KS_sf, size = 0.4) +
  geom_sf(data = KS_county, fill = NA)
```

```{r}
# CRS
st_crs(KS_county)
```

```{r}
# Convert the CRS to WGS 84/UTM zone 14N (EPSG code: 32614)
g_32614 <- st_transform(KS_county, 32614) %>%
  ggplot() +
  geom_sf()

g_county
g_32614

# Alternatively
KS_county %>%
  ggplot() +
  geom_sf() +
  coord_sf(crs = 32614)
```

```{r}
# CRS of the first layer is applied for all the layers
ggplot() +
  geom_sf(
    data = st_transform(KS_county, 32614)
  ) +
  geom_sf(data = gw_KS_sf, size = 0.4)
```

```{r}
# Limit the geographic scope of the map
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, size = 0.4) +
  coord_sf(crs = 4269) +
  xlim(-99, -97) +
  ylim(37, 39)
```

```{r}
# Faceting
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, aes(color = af_used)) +
  facet_wrap(vars(year))
```

```{r}
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, aes(color = af_used)) +
  facet_wrap(
    vars(year),
    ncol = 1
  )
```

```{r}
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, aes(color = af_used)) +
  facet_wrap((af_used > 200) ~ year) +
  theme(panel.spacing.y = unit(1, "lines"))
```

```{r}
# Adding texts (labels) on a map
p <- ggplot() +
  geom_sf(data = KS_county) 

p +
  geom_sf_text(
    data      = KS_county,
    aes(label = NAME),
    size      = 3,
    color     = "blue"
  )

p +
  geom_sf_text(
    data          = KS_county,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue"
  )

p +
  geom_sf_text(
    data          = KS_county,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue",
    nudge_x       = -0.1,
    nudge_y       = 0.1
  )
```

```{r}
# Individually
Cheyenne <- KS_county %>% 
  filter(NAME == "Cheyenne")

KS_less_Cheyenne <- KS_county %>% 
  filter(NAME != "Cheyenne")

p <- ggplot() +
  geom_sf(data = KS_county) +
  geom_sf_text(
    data          = KS_less_Cheyenne,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue",
    nudge_x       = -0.1,
    nudge_y       = 0.1
  ) 

p +
  geom_sf_text(
    data          = Cheyenne,
    aes(label     = NAME),
    size          = 2.5,
    color         = "red",
    nudge_y       = 0.2
  )

# With annotate()
p +
  annotate(
    geom  = "text",
    x     = -102,
    y     = 39.8,
    size  = 3,
    label = "Cheyennes",
    color = "red"
  )
```

#### 3.7.3.2 Arranging maps

```{r}
# Inset
three_counties <- KS_county %>% 
  filter(
    NAME %in% c("Cheyenne", "Sherman", "Wallace")
  )

g_three_counties <- three_counties %>% 
  ggplot() +
  geom_sf() +
  geom_sf_text(aes(label = NAME)) +
  theme_void()

g_three_counties
```

```{r}
# Convert the ggplot into a grob
grob_aoi <- ggplotGrob(g_three_counties)

# Check the class
class(grob_aoi)

# Create a map of Kansas
g_region <- ggplot() +
  geom_sf(data = KS_county) +
  geom_sf(
    data  = three_counties,
    fill  = "blue",
    color = "red",
    alpha = 0.5) +
  theme_void()

# Convert to a grob
grob_region <- ggplotGrob(g_region)

g_inset <- ggplot() +
  coord_equal(
    xlim   = c(0, 1),
    ylim   = c(0, 1),
    expand = FALSE
  )

g_inset
```

```{r}
g_aoi <- g_inset +
  annotation_custom(
    grob_aoi,
    xmin = 0,
    xmax = 0.8,
    ymin = 0,
    ymax = 1
  )

g_aoi

g_region <- g_aoi +
  annotation_custom(
    grob_region,
    xmin = 0.6,
    xmax = 1,
    ymin = 0,
    ymax = 0.3
  )

g_region
```

```{r}
# Without border
g_region +
  theme_void()
```

#### 3.7.3.3 Fine-tuning maps for publication

```{r}
gw_by_county <- st_join(
  KS_county,
  gw_KS_sf
) %>%
  as_tibble() %>%
  summarize(
    af_used = sum(af_used, na.rm = TRUE),
    .by     = c(COUNTYFP, year)
  ) %>% 
  left_join(KS_county, ., by = "COUNTYFP") %>%
  filter(!is.na(year))

gw_by_county
```

```{r}
g_base <- gw_by_county %>% 
  ggplot() +
  geom_sf(aes(fill = af_used)) +
    facet_wrap(vars(year))

g_base
```

```{r}
# Setting the theme
g_base +
  theme_void()

theme_void()
```

```{r}
g_base +
  theme(axis.text = element_blank())
```

```{r}
theme_for_map <-
  theme(
    axis.ticks       = element_blank(),
    axis.text        = element_blank(),
    axis.line        = element_blank(),
    panel.border     = element_blank(),
    panel.grid       = element_line(
      color = "transparent"),
    panel.background = element_blank(),
    plot.background  = element_rect(
      fill  = "transparent",
      color = "transparent")
  )

g_base +
  theme_for_map
```

```{r}
theme_for_map_2 <- 
  theme_void() +
  theme(
    strip.background = element_rect(
      fill = "gray"
    )
  )

g_base +
  theme_for_map_2
```

```{r}
p <- gw_by_county %>%
  ggplot(aes(fill = af_used)) +
  geom_sf() +
  facet_grid(vars(year)) +
  theme_bw()

p

p +
  theme(
    panel.grid = element_line(
      color = "transparent"
    )
  )
```

```{r}
g_legend <- gw_by_county %>% 
  ggplot(aes(fill = af_used)) +
  geom_sf() +
  facet_wrap(vars(year)) +
  theme_void()

g_legend +
  scale_fill_viridis_c() +
  labs(
    fill = "Groundwater pumping (acre-feet)"
  ) +
  theme(
    legend.position   = "bottom",
    legend.key.height = unit(0.5, "cm"),
    legend.key.width  = unit(2, "cm"),
    legend.text       = element_text(
      size   = 12,
      family = "Times"
    ),
    legend.title      = element_text(
      size   = 12,
      family = "Times"
    )
  ) +
  guides(
    fill = guide_colorbar(title.position = "top")
  )
```

### 3.7.4 Area Metadata

```{r}
mi_census <- midwest %>%
  filter(state == "MI") %>%
  # tolower -> str_to_lower
  mutate(county = str_to_lower(county)) %>%
  select(
    county,
    area,
    poptotal,
    percwhite,
    percblack
  )

mi_census
```

```{r}
census_counties <- mi_census %>% 
  left_join(
    mi_counties,
    by = c("county" = "id")
  )

census_counties
```

```{r}
p <- census_counties %>% 
  ggplot(aes(lon, lat, group = county))
  
p +
  geom_polygon(aes(fill = poptotal)) +
  coord_sf()

p +
  geom_polygon(aes(fill = percwhite)) +
  coord_sf()
```

## 3.8 Revealing Uncertainty

```{r}
y   <- c(18, 11, 16)
tbl <- tibble(
  x  = 1:3,
  y  = y,
  se = c(1.2, 0.5, 1.0)
)

base <- tbl %>% 
  ggplot(aes(x, y, ymin = y - se, ymax = y + se))

# Added the titles, so it's easier to remember which one you're looking at
base +
  geom_crossbar() +
  labs(title = "geom_crossbar")

base +
  geom_pointrange() +
  labs(title = "geom_pointrange")

base +
  geom_smooth(stat = "identity") +
  labs(title = "geom_smooth")
  
base +
  geom_errorbar() +
  labs(title = "geom_errorbar")

base +
  geom_linerange() +
  labs(title = "geom_linerange")

base +
  geom_ribbon() +
  labs(title = "geom_ribbon")
```

## 3.9 Weighted Data

```{r}
# Unweighted
p <- midwest %>%
  ggplot(aes(percwhite, percbelowpoverty))

p +
  geom_point()

# Weight by population
p + 
  geom_point(aes(size = poptotal / 1e6)) + 
  scale_size_area(
    "Population\n(millions)",
    breaks = c(0.5, 1, 2, 4)
  )
```

```{r}
# Unweighted
p +
  geom_point() + 
  geom_smooth(method = lm, linewidth = 1)

# Weighted by population
p + 
  geom_point(aes(size = poptotal / 1e6)) + 
  geom_smooth(
    aes(weight = poptotal),
    method = lm,
    linewidth = 1
  ) +
  scale_size_area(guide = "none")
```

```{r}
p <- midwest %>%
  ggplot(aes(percbelowpoverty))

p +
  geom_histogram(binwidth = 1) + 
  labs(y = "Counties")

p +
  geom_histogram(
    aes(weight = poptotal),
    binwidth = 1
  ) +
  labs(y = "Population (1000s)")
```

## 3.10 Diamonds Data

```{r}
diamonds
```

## 3.11 Displaying Distributions

```{r}
p <- diamonds %>%
  ggplot(aes(depth))

p + 
  geom_histogram()

p +
  geom_histogram(binwidth = 0.1) + 
  xlim(55, 70)
```

```{r}
p <- diamonds %>%
  ggplot(aes(depth))

p + 
  geom_freqpoly(
    aes(color = cut),
    binwidth   = 0.1,
    na.rm      = TRUE
  ) +
  xlim(58, 68) + 
  theme(legend.position = "none")

p + 
  geom_histogram(
    aes(fill = cut),
    binwidth = 0.1,
    position = "fill",
    na.rm    = TRUE
  ) +
  xlim(58, 68) + 
  theme(legend.position = "none")
```

```{r}
diamonds %>%
  ggplot(aes(depth)) +
  geom_density(na.rm = TRUE) +
  xlim(58, 68) +
  theme(legend.position = "none")

diamonds %>%
  ggplot(aes(depth, fill = cut, color = cut)) +
  geom_density(alpha = 0.2, na.rm = TRUE) +
  xlim(58, 68) +
  theme(legend.position = "none")
```

```{r}
diamonds %>%
  ggplot(aes(clarity, depth)) +
  geom_boxplot()

diamonds %>%
  ggplot(aes(carat, depth)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1))) +
  xlim(NA, 2.05)
```

```{r}
diamonds %>%
  ggplot(aes(clarity, depth)) + 
  geom_violin()

diamonds %>%
  ggplot(aes(carat, depth)) + 
  geom_violin(aes(group = cut_width(carat, 0.1))) + 
  xlim(NA, 2.05)
```

### 3.11.1 Exercises

1.
```{r}
diamonds %>% 
  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.2)
```

2.
```{r}
p <- diamonds %>% 
  ggplot(aes(price))

p +
  geom_histogram(binwidth = 100)

p +
  geom_histogram(binwidth = 500)
```

3.
```{r}
diamonds %>% 
  ggplot(aes(clarity, price)) +
  geom_boxplot()
```

4. 
```{r}
diamonds %>% 
  count(depth) %>% 
  mutate(
    sum     = sum(n),
    density = n / sum
  ) %>% 
  ggplot(aes(depth, density)) +
  geom_line()
```

## 3.12 Dealing with Overplotting

```{r}
tbl <- tibble(x = rnorm(2000), y = rnorm(2000))

norm <- tbl %>% 
  ggplot(aes(x, y)) +
  labs(
    x = NULL,
    y = NULL
  )

# Added the titles, so it's easier to remember which one you're looking at

norm +
  geom_point()

# Hollow circles
norm +
  geom_point(shape = 1) +
  labs(title = "shape = 1")

# Pixel sized
norm +
  geom_point(shape = ".") +
  labs(title = 'shape = "."')

norm +
  geom_point(alpha = 1 / 3) +
  labs(title = "alpha = 1 / 3")

norm +
  geom_point(alpha = 1 / 5) +
  labs(title = "alpha = 1 / 5")

norm +
  geom_point(alpha = 1 / 10) +
  labs(title = "alpha = 1 / 10")

```

```{r}
norm +
  geom_bin2d()

norm +
  geom_bin2d(bins = 10)
```

```{r}
norm +
  geom_hex()

norm +
  geom_hex(bins = 10)
```

## 3.13. Statistical Summaries

```{r}
diamonds %>%
  ggplot(aes(color)) + 
  geom_bar()

diamonds %>%
  ggplot(aes(color, price)) + 
  geom_bar(stat = "summary_bin", fun = mean)
```

```{r}
diamonds %>%
  ggplot(aes(table, depth)) + 
  geom_bin2d(binwidth = 1, na.rm = TRUE) + 
  xlim(50, 70) + 
  ylim(50, 70)

diamonds %>%
  ggplot(aes(table, depth, z = price)) + 
  geom_raster(
    binwidth = 1,
    stat = "summary_2d",
    fun = mean, 
    na.rm = TRUE
  ) + 
  xlim(50, 70) + 
  ylim(50, 70)

diamonds %>%
  ggplot(aes(table, depth, z = price)) + 
  geom_tile(
    binwidth = 1,
    stat = "summary_2d",
    fun = mean, 
    na.rm = TRUE
  ) + 
  xlim(50, 70) + 
  ylim(50, 70)
```

## 3.X

### rgl - 3D visualization device system for R using OpenGL

https://dmurdoch.github.io/rgl/

```{r}
library(rgl)

with(
  iris,
  plot3d(
    Sepal.Length,
    Sepal.Width,
    Petal.Length, 
    type = "s",
    col  = as.numeric(Species)
  )
)
```

```{r}
close3d()
```

```{r}
library(rgl)
library(MASS)

set.seed(123)
x   <- rgamma(100, shape = 5, rate = 0.1)
fit <- fitdistr(x, dgamma, list(shape = 1, rate = 0.1), lower = 0.001)

loglik <- function(shape, rate) {
  sum(dgamma(x, shape = shape, rate = rate, log = TRUE)) 
}

loglik <- Vectorize(loglik)
xlim <- fit$estimate[1] + 4 * fit$sd[1] * c(-1,1)
ylim <- fit$estimate[2] + 4 * fit$sd[2] * c(-1,1)
mfrow3d(1, 2, sharedMouse = TRUE)
persp3d(
  loglik,
  xlim = xlim,
  ylim = ylim,
  n = 30
)
zlim <- fit$loglik + c(-qchisq(0.99, 2)/2, 0)
next3d()
persp3d(
  loglik, 
  xlim = xlim,
  ylim = ylim,
  zlim = zlim,
  n = 30
)
```

```{r}
close3d()
```

```{r}
library(rgl)
triangles3d(
  cbind(
    x = rnorm(9),
    y = rnorm(9),
    z = rnorm(9)
  ),
  col = "green"
)
decorate3d()
bg3d("lightgray")
aspect3d(1, 1, 1)
```

```{r}
close3d()
```

### OpenStreetMap with Dr. Dominic Royé

https://dominicroye.github.io/en/2018/accessing-openstreetmap-data-with-r/

```{r}
library(osmdata)
library(sf)
library(ggmap)

# The first five features
available_features() %>% 
  head()

# Amenities
available_tags("amenity") %>% 
  head()

# Shops
available_tags("shop") %>% 
  head()
```

```{r}
# Building the query
q <- getbb("Madrid") %>%
  opq() %>%
  add_osm_feature("amenity", "cinema")

# Query structure
q %>% 
str() 
```

```{r}
cinema <- osmdata_sf(q)
cinema
```

```{r}
# Our background map
mad_map <- get_map(
  getbb("Madrid"),
  maptype = "toner-background",
  source  = "stamen"
)

# Final map
mad_map %>%
  ggmap() +
  geom_sf(
    data        = cinema$osm_points,
    inherit.aes = FALSE,
    color       = "#238443",
    fill        = "#004529",
    alpha       = .5,
    size        = 4,
    shape       = 21
  ) +
  labs(
    x = "",
    y = ""
  )
```

```{r}
# Bounding box for the Iberian Peninsula
m <- c(-10, 30, 5, 46)

# Building the query
q <- m %>% 
  opq(timeout = 25*100) %>%
  add_osm_feature("name", "Mercadona") %>%
  add_osm_feature("shop", "supermarket")

# Query
mercadona <- osmdata_sf(q)

# Final map
ggplot(mercadona$osm_points) +
  geom_sf(
    color = "#08519c",
    fill  = "#08306b",
    alpha = .5,
    size  = 1,
    shape = 21
    ) +
  theme_void()
```
