---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (3rd ed.), Chapter 9 - Arranging plots"
author: "Original Code: Hadley Wickham (except Thomas Lin Pedersen for patchwork, Claus O. Wilke for cowplot, and Alboukadel Kassambara for ggpubr) | Modifications: Antti Rask"
date: "2023-01-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 9 Arranging plots

```{r}
library(patchwork)
library(tidyverse)
```

## 9.1 Laying out plots side by side

```{r}
p1 <- mpg %>%
  ggplot(aes(displ, hwy)) + 
  geom_point() +
  labs(
    title = "Plot 1"
  )

p2 <- mpg %>%
  ggplot(aes(as.character(year), fill = drv)) + 
  geom_bar(position = "dodge") + 
  labs(
    title = "Plot 2",
    x = "year"
  )

p3 <- mpg %>%
  ggplot(aes(hwy, fill = drv)) + 
  geom_density(colour = NA) + 
  facet_grid(rows = vars(drv)) +
  labs(
    title = "Plot 3"
  )

p4 <- mpg %>% 
  ggplot(aes(drv, hwy)) + 
  stat_summary(
    aes(fill = drv),
    geom     = "col",
    fun.data = mean_se
  ) +
  stat_summary(
    geom     = "errorbar",
    fun.data = mean_se,
    width    = 0.5
  ) +
  labs(
    title = "Plot 4"
  )

p1 + p2

p1 + p2 + p3

p1 + p2 + p3 + p4
```

### 9.1.1 Taking control of the layout

```{r}
p1 + p2 + p3 +
  plot_layout(ncol = 2)

p1 / p2

# Basically the same as using `+` but the intent is clearer
p3 | p4

p3 | (p2 / (p1 | p4))
```

```{r}
layout <- 
  "
AAB
C#B
CDD
"

p1 + p2 + p3 + p4 +
  plot_layout(design = layout)
```

```{r, include=FALSE}
p <- p1 + p2 + p3 + 
  plot_layout(ncol = 2, guides = "collect")
```

```{r}
p

p +
  guide_area()
```


### 9.1.2 Modifying subplots

```{r, include=FALSE}
p12 <- p1 + p2

p12[[2]] <- p12[[2]] + 
  theme_light()
```

```{r}
p12

p1 + p4 &
  theme_minimal()

p1 + p4 &
  scale_y_continuous(limits = c(0, 45))

# * will add the element to all the subplots in the current nesting level
p1 / (p3 | p4) *
  theme_minimal()
```

### 9.1.3 Adding annotation

```{r, include=FALSE}
p34 <- p3 + p4 + 
  plot_annotation(
    title = "A closer look at the effect of drive train in cars",
    caption = "Source: mpg dataset in ggplot2"
  )
```

```{r}
p34

p34 + 
  plot_annotation(theme = theme_gray(base_family = "mono"))

p34 & 
  theme_gray(base_family = "mono")
```

```{r, include=FALSE}
p123 <- p1 | (p2 / p3)
```

```{r}
# Uppercase roman numerals
p123 + 
  plot_annotation(tag_levels = "I") 
```

```{r, include=FALSE}
p123[[2]] <- p123[[2]] + 
  plot_layout(tag_level = "new")
```

```{r}
p123 + 
  plot_annotation(tag_levels = c("I", "a"))
```

## 9.2 Arranging plots on top of each other

```{r}
p1 + 
  inset_element(
    p2,
    left   = 0.5,
    bottom = 0.4,
    right  = 0.9,
    top    = 0.95)

p1 + 
  inset_element(
    p2, 
    left     = 0.4, 
    bottom   = 0.4, 
    right    = unit(1, "npc") - unit(15, "mm"), 
    top      = unit(1, "npc") - unit(15, "mm"),
    align_to = "full"
  )
```

```{r, include=FALSE}
p24 <- p2 / p4 +
  plot_layout(guides = "collect")
```

```{r}
p1 +
  inset_element(
    p24,
    left   = 0.5,
    bottom = 0.05,
    right  = 0.95,
    top    = 0.9
  )
```

```{r, include=FALSE}
p12 <- p1 +
  inset_element(
    p2,
    left   = 0.5,
    bottom = 0.5,
    right  = 0.9,
    top    = 0.95
  )
```

```{r}
p12 &
  theme_bw()

p12 +
  plot_annotation(tag_levels = "A")
```

## 9.X

### cowplot - Streamlined Plot Theme and Plot Annotations for ggplot2

Original from: https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html

#### Themes

```{r}
library(cowplot)

# the plot
p <- iris %>% 
  ggplot(aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point()

# default ggplot2 theme
p

# classic cowplot theme
p +
  theme_cowplot(12)

# minimal grid theme
p +
  theme_minimal_grid(12)

# minimal horizontal grid theme
p2 <- iris %>%
  ggplot(aes(Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

p2 +
  theme_minimal_hgrid(12)



```

#### Arranging plots into a grid

```{r}
p1 <- mtcars %>%
  ggplot(aes(disp, mpg)) +
  geom_point()

p2 <- mtcars %>%
  ggplot(aes(qsec, mpg)) +
  geom_point()

plot_grid(p1, p2, labels = c('A', 'B'), label_size = 12)

p3 <- ~plot(mtcars$qsec, mtcars$mpg)

plot_grid(p1, p3, labels = c('A', 'B'), label_size = 12)
```

#### Generic plot annotations

```{r}
p4 <- mtcars %>% 
  ggplot(aes(disp, mpg)) +
  geom_point(size = 1.5, color = "blue") +
  theme_cowplot(12)

logo_file <- system.file("extdata", "logo.png", package = "cowplot")

ggdraw(p4) +
  draw_image(
    logo_file,
    x     = 1,
    y     = 1,
    hjust = 1,
    vjust = 1,
    width = 0.13,
    height = 0.2
  )

ggdraw() + 
  draw_image(logo_file, scale = 0.5) +
  draw_plot(p4)
```

### ggpubr - ‘ggplot2’ Based Publication Ready Plots

Original from: https://rpkgs.datanovia.com/ggpubr/

#### Distribution

```{r}
library(ggpubr)

set.seed(1234)

wdata <- tibble(
  sex    = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))

wdata %>% 
  ggdensity(
    x       = "weight",
    add     = "mean",
    rug     = TRUE,
    color   = "sex",
    fill    = "sex",
    palette = c("#00AFBB", "#E7B800")
  )

wdata %>% 
  gghistogram(
    x       = "weight",
    add     = "mean",
    rug     = TRUE,
    color   = "sex",
    fill    = "sex",
    palette = c("#00AFBB", "#E7B800")
  )
```

#### Box and violin plots

```{r}
tbl <- ToothGrowth

p <- tbl %>% 
  ggboxplot(
    x       = "dose",
    y       = "len",
    color   = "dose",
    palette = c("#00AFBB", "#E7B800", "#FC4E07"),
    add     = "jitter",
    shape   = "dose"
  )

p

# Add p-values comparing groups
my_comparisons <- list( c("0.5", "1"), c("1", "2"), c("0.5", "2") )

p +
  # Add pairwise comparisons p-value
  stat_compare_means(comparisons = my_comparisons) + 
  # Add global p-value
  stat_compare_means(label.y = 50)
```

```{r}
tbl %>% 
  ggviolin(
    x         = "dose",
    y         = "len",
    fill      = "dose",
    palette   = c("#00AFBB", "#E7B800", "#FC4E07"),
    add       = "boxplot", 
    add.params = list(fill = "white")
  ) +
  # Add significance levels
  stat_compare_means(comparisons = my_comparisons, label = "p.signif") + 
  # Add global the p-value
  stat_compare_means(label.y = 50)
```
#### Bar plots

```{r}
tbl <- mtcars %>% 
  as_tibble(rownames = "rowname") %>% 
  rename("name" = "rowname") %>% 
  mutate(cyl = as.factor(cyl))

tbl
```

```{r}
tbl %>% 
  ggbarplot(
    x              = "name",
    y              = "mpg",
    fill           = "cyl",
    color          = "white",
    palette        = "jco",         
    sort.val       = "asc",
    sort.by.groups = FALSE,
    rotate         = TRUE
  )

tbl %>% 
  ggbarplot(
    x              = "name",
    y              = "mpg",
    fill           = "cyl",
    color          = "white",
    palette        = "jco",         
    sort.val       = "asc",
    sort.by.groups = TRUE,
    rotate         = TRUE
  )
```

```{r}
# Calculate the z-score of the mpg data
tbl2 <- tbl %>%
  mutate(
    mpg_z   = (mpg - mean(mpg)) / sd(mpg),
    mpg_grp = case_when(
      mpg_z < 0 ~ "low",
      TRUE      ~ "high"
    ) %>% 
      as_factor() %>%
      fct_relevel("low", "high")
  )

tbl2
```

```{r}
tbl2 %>% 
  ggbarplot(
    x              = "name",
    y              = "mpg_z",
    fill           = "mpg_grp",
    color          = "white",
    palette        = "jco",       
    sort.val       = "asc",
    sort.by.groups = FALSE,
    ylab           = "MPG z-score",
    xlab           = FALSE,
    legend.title   = "MPG Group",
    rotate         = TRUE
  )

tbl2 %>% 
  ggbarplot(
    x              = "name",
    y              = "mpg_z",
    fill           = "mpg_grp",
    color          = "white",
    palette        = "jco",       
    sort.val       = "asc",
    sort.by.groups = FALSE,
    ylab           = "MPG z-score",
    xlab           = FALSE,
    legend.title   = "MPG Group",
    rotate         = TRUE,
    ggtheme        = theme_minimal()
  )
```

#### Dot charts

```{r}
tbl %>% 
  ggdotchart(
    x              = "name",
    y              = "mpg",
    color          = "cyl",
    palette        = c("#00AFBB", "#E7B800", "#FC4E07"),
    sorting        = "descending",
    add            = "segments",  
    ggtheme        = theme_pubr(),
    rotate         = TRUE
  )

tbl %>% 
  ggdotchart(
    x              = "name",
    y              = "mpg",
    color          = "cyl",
    palette        = c("#00AFBB", "#E7B800", "#FC4E07"),
    sorting        = "descending",
    add            = "segments",
    rotate         = TRUE,
    group          = "cyl",
    dot.size       = 6,
    label          = round(tbl$mpg),
    font.label     = list(color = "white", size = 9, vjust = 0.5),
    ggtheme        = theme_pubr()
  )

tbl2 %>% 
  ggdotchart(
    x              = "name",
    y              = "mpg_z",
    color          = "cyl",
    palette        = c("#00AFBB", "#E7B800", "#FC4E07"),
    sorting        = "ascending",
    add            = "segments",
    add.params     = list(color = "lightgray", size = 2),
    rotate         = TRUE,
    group          = "cyl",
    dot.size       = 8,
    label          = round(tbl2$mpg_z, 1),
    font.label     = list(color = "white", size = 9, vjust = 0.5),
    ggtheme        = theme_pubr()
  ) +
  geom_hline(yintercept = 0, linetype = 2, color = "lightgray")

tbl %>%
  ggdotchart(
    x              = "name",
    y              = "mpg",
    color          = "cyl",
    palette        = c("#00AFBB", "#E7B800", "#FC4E07"),
    sorting        = "descending",
    rotate         = TRUE,
    dot.size       = 2,
    y.text.col     = TRUE,
    ggtheme        = theme_pubr()
  ) +
  theme_cleveland()

```

### patchwork

Original from: https://patchwork.data-imaginist.com/

#### Adding non-ggplot content

```{r}
library(grid)
library(gridExtra)

p1 + 
  textGrob('Some really important text')


wrap_elements(textGrob('Text on left side')) + p1

p1 +
  tableGrob(mtcars[1:10, c('mpg', 'disp')])
```

```{r}
# 1
p1 + 
  ~plot(mtcars$mpg, mtcars$disp, main = 'Plot 2')

# 2
old_par <- par(mar = c(0, 2, 0, 0), bg = NA)

p1 +
  wrap_elements(panel = ~plot(mtcars$mpg, mtcars$disp), clip = FALSE)

par(old_par)

old_par <- par(mar = c(0, 0, 0, 0), mgp = c(1, 0.25, 0), 
               bg = NA, cex.axis = 0.75, las = 1, tcl = -0.25)

# 3
p1 + 
  wrap_elements(panel = ~plot(mtcars$mpg, mtcars$disp), clip = FALSE) +
  ggtitle('Plot 2') +
  theme(plot.margin = margin(5.5, 5.5, 5.5, 35))

par(old_par)
```

#### Adding an empty area

```{r}
p1 + 
  plot_spacer() +
  p2 + 
  plot_spacer() + 
  p3 + 
  plot_spacer()

# Spacers in separate nesting levels may have different dimensions
(p1 + plot_spacer() + p2) / (plot_spacer() + p3 + plot_spacer())
``` 

#### Moving beyond the grid

```{r}
layout <- c(
  area(t = 2, l = 1, b = 5, r = 4),
  area(t = 1, l = 3, b = 3, r = 5)
)

p1 + p2 + 
  plot_layout(design = layout)
```

```{r}
layout <- '
A#B
#C#
D#E
'
wrap_plots(
  D      = p1,
  C      = p2,
  B      = p3,
  design = layout
)
```

#### Fixed aspect plots

```{r}
p_fixed <- mtcars %>% 
  ggplot(aes(hp, disp)) + 
  geom_point() + 
  ggtitle("Plot F") + 
  coord_fixed()

p_fixed + p1 + p2 + p3

p_fixed + p1 + p2 + p3 +
  plot_layout(widths = 1)
```

#### Alignment across multiple pages

```{r, include = FALSE}
p3_dims    <- get_dim(p3)
p1_aligned <- set_dim(p1, p3_dims)
```

```{r}
plot(p1_aligned)

p1_aligned +
  theme_minimal()

plot(p3)
```

```{r, include=FALSE}
max_dims <- get_max_dim(p1, p2, p3, p4)
```

```{r}
set_dim(p4, max_dims)

set_dim(p2, max_dims)
```

```{r, include=FALSE}
plots_aligned <- align_patches(p1, p2, p3, p4)
```

```{r}
for (p in plots_aligned) {
  plot(p)
}
```
