---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (3rd ed.), Chapter 6 - Maps"
author: "Original Code: Hadley Wickham (except Taro Mieno for Raster Images, Duncan Murdoch for rgl and Dr. Dominic Royé for OpenStreetMaps) | Modifications: Antti Rask"
date: "2023-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 6 Maps

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 6.1 Polygon maps

```{r}
mi_counties <- map_data("county", "michigan") %>%
  as_tibble() %>%
  select(lon = long, lat, group, id = subregion)
  
head(mi_counties)
```

```{r}
p <- mi_counties %>%
  ggplot(aes(lon, lat, group = group))

p +
  geom_point(size = .25, show.legend = FALSE) +
  coord_sf()
  
p +
  geom_polygon() +
  coord_sf()

p +
  geom_polygon(
    fill = NA, 
    color = "grey50"
  ) + 
  coord_sf()
```

## 6.2 Simple features map

```{r}
library(ozmaps)
library(sf)

oz_states <- ozmap_states
oz_states

oz_states %>%
  ggplot() + 
  geom_sf() + 
  coord_sf()
```

### 6.2.1 Layered maps

```{r}
library(ozmaps)
library(rmapshaper)

oz_states <- ozmap_states %>%
  filter(NAME != "Other Territories")

oz_votes <- ms_simplify(abs_ced)

ggplot() + 
  geom_sf(
    data        = oz_states,
    mapping     = aes(fill = NAME),
    show.legend = FALSE
  ) +
  geom_sf(data = oz_votes, fill = NA) + 
  coord_sf()
```

### 6.2.2 Labelled maps

```{r}
library(ozmaps)
# Filter electorates in the Sydney metropolitan region
sydney_map <- abs_ced %>%
  filter(
    NAME %in% c(
      "Sydney",
      "Wentworth",
      "Warringah",
      "Kingsford Smith",
      "Grayndler",
      "Lowe",
      "North Sydney",
      "Barton",
      "Bradfield",
      "Banks",
      "Blaxland",
      "Reid",
      "Watson",
      "Fowler",
      "Werriwa",
      "Prospect",
      "Parramatta",
      "Bennelong",
      "Mackellar",
      "Greenway",
      "Mitchell",
      "Chifley",
      "McMahon"
    )
  )

# Draw the electoral map of Sydney
sydney_map %>%
ggplot() +
  geom_sf(aes(fill = NAME), show.legend = FALSE) + 
  coord_sf(
    xlim = c(150.97, 151.3),
    ylim = c(-33.98, -33.79)
  ) + 
  geom_sf_label(
    aes(label     = NAME),
    label.padding = unit(1, "mm")
  )

# From the book:
# "The warning message is worth noting. Internally geom_sf_label() uses the function st_point_on_surface() from the sf package to place labels, and the warning message occurs because most algorithms used by sf to compute geometric quantities (e.g., centroids, interior points) are based on an assumption that the points lie in on a flat two dimensional surface and parameterised with Cartesian co-ordinates. This assumption is not strictly warranted, and in some cases (e.g., regions near the poles) calculations that treat longitude and latitude in this way will give erroneous answers. For this reason, the sf package produces warning messages when it relies on this approximation."
```

### 6.2.3 Adding other geoms

```{r}
oz_capitals <- tribble( 
  ~city,           ~lat,     ~lon,
  "Sydney",    -33.8688, 151.2093,  
  "Melbourne", -37.8136, 144.9631, 
  "Brisbane",  -27.4698, 153.0251, 
  "Adelaide",  -34.9285, 138.6007, 
  "Perth",     -31.9505, 115.8605, 
  "Hobart",    -42.8821, 147.3272, 
  "Canberra",  -35.2809, 149.1300, 
  "Darwin",    -12.4634, 130.8456, 
)

ggplot() + 
  geom_sf(data = oz_votes) + 
  geom_sf(
    data  = oz_states,
    color = "black",
    fill  = NA
  ) + 
  geom_point(
    data    = oz_capitals,
    mapping = aes(x = lon, y = lat), color = "red") + 
  coord_sf()
```

## 6.3 Map projections

```{r}
st_crs(oz_votes)

st_crs(oz_votes) == st_crs(4283)
```

```{r}
p <- oz_votes %>%
  ggplot() +
  geom_sf()

p

p +
  coord_sf(crs = st_crs(3112))
```

## 6.4 Working with sf data

```{r}
library(ozmaps)

edenmonaro <- abs_ced %>%
  filter(NAME == "Eden-Monaro")

p <- edenmonaro %>% 
  ggplot() +
  geom_sf()

p + 
  coord_sf(
    xlim = c(147.75, 150.25),
    ylim = c(-37.5, -34.5)
  ) 

p +
  coord_sf(
    xlim = c(150, 150.25),
    ylim = c(-36.3, -36)
  ) 
```

```{r}
edenmonaro_geometry <- edenmonaro %>% 
  pull(geometry)

st_bbox(edenmonaro_geometry)
```

```{r}
edenmonaro_geometry
```

```{r}
st_cast(edenmonaro_geometry, "POLYGON")
```

```{r}
library(ozmaps)
dawson <- abs_ced %>% 
  filter(NAME == "Dawson") %>% 
  pull(geometry)

dawson

dawson %>% 
  ggplot() + 
  geom_sf() +
  coord_sf()
```

```{r}
dawson <- st_cast(dawson, "POLYGON")

which.max(st_area(dawson))

dawson[-69] %>%
  ggplot() + 
  geom_sf() + 
  coord_sf()
```

## 6.5 Raster maps

There was a problem with the {bomrang} package and it has been archived. So, the example code for the 3rd
edition doesn't work as it is.

I opened an issue about it (https://github.com/hadley/ggplot2-book/issues/338#issuecomment-1422203499), but so far there is no idea how Hadley & co. are going to replace the chapter.

While waiting for the update, I'll include another example: https://tmieno2.github.io/R-as-GIS-for-Economists/create-maps.html

### Creating maps from sf objects

```{r}
# Read in the KS wells data
gw_KS_sf <- readRDS("input/gw_KS_sf.rds")
gw_KS_sf
```

```{r}
# Polygons
library(sf)
library(tigris)

KS_county <- counties(
  state = "Kansas",
  cb    = TRUE
) %>%
  st_as_sf() %>%
  st_transform(st_crs(gw_KS_sf))

KS_county
```

```{r}
# Points
g_wells <- gw_KS_sf %>% 
  ggplot() +
  geom_sf()

g_wells
```

```{r}
g_county <- KS_county %>%
  ggplot() +
  geom_sf()

g_county
```

```{r}
# Points
gw_Stevens <- KS_county %>%
  filter(NAME == "Stevens") %>%
  st_crop(gw_KS_sf, .)

p <- 
  gw_Stevens %>% 
  ggplot()

p +
  geom_sf(aes(color = af_used), size = 2)

p +
  geom_sf(
    aes(size = af_used),
    color    = "blue",
    shape    = 15
  )

gw_Stevens %>%
  cbind(., st_coordinates(.)) %>%
   mutate(
     east_west = ifelse(
       X < -101.3,
       "west",
       "east")
   ) %>%
  ggplot() +
  geom_sf(
    aes(shape = east_west, color = east_west)
  )
```

```{r}
# Polygons
KS_county %>%
  ggplot() +
  geom_sf(color = "red", fill = "darkgreen")

KS_county_with_pumping <- gw_KS_sf %>%
  filter(year == 2010) %>%
  aggregate(., KS_county, sum, na.rm = TRUE)

KS_county_with_pumping %>%
  ggplot() +
  geom_sf(aes(fill = af_used))
```

```{r}
# Plotting multiple spatial objects in one figure
ggplot() +
  geom_sf(data = gw_KS_sf, size = 0.4) +
  geom_sf(data = KS_county, fill = NA)
```

```{r}
# CRS
st_crs(KS_county)
```

```{r}
# Convert the CRS to WGS 84/UTM zone 14N (EPSG code: 32614)
g_32614 <- st_transform(KS_county, 32614) %>%
  ggplot() +
  geom_sf()

g_county
g_32614

# Alternatively
KS_county %>%
  ggplot() +
  geom_sf() +
  coord_sf(crs = 32614)
```

```{r}
# CRS of the first layer is applied for all the layers
ggplot() +
  geom_sf(
    data = st_transform(KS_county, 32614)
  ) +
  geom_sf(data = gw_KS_sf, size = 0.4)
```

```{r}
# Limit the geographic scope of the map
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, size = 0.4) +
  coord_sf(crs = 4269) +
  xlim(-99, -97) +
  ylim(37, 39)
```

```{r}
# Faceting
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, aes(color = af_used)) +
  facet_wrap(vars(year))
```

```{r}
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, aes(color = af_used)) +
  facet_wrap(
    vars(year),
    ncol = 1
  )
```

```{r}
ggplot() +
  geom_sf(data = st_transform(KS_county, 32614)) +
  geom_sf(data = gw_KS_sf, aes(color = af_used)) +
  facet_wrap((af_used > 200) ~ year) +
  theme(panel.spacing.y = unit(1, "lines"))
```

```{r}
# Adding texts (labels) on a map
ggplot() +
  geom_sf(data = KS_county) +
  geom_sf_text(
    data      = KS_county,
    aes(label = NAME),
    size      = 3,
    color     = "blue"
  )

ggplot() +
  geom_sf(data = KS_county) +
  geom_sf_text(
    data          = KS_county,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue"
  )

ggplot() +
  geom_sf(data = KS_county) +
  geom_sf_text(
    data          = KS_county,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue",
    nudge_x       = -0.1,
    nudge_y       = 0.1
  )
```

```{r}
# Individually
Cheyenne <- KS_county %>% 
  filter(NAME == "Cheyenne")

KS_less_Cheyenne <- KS_county %>% 
  filter(NAME != "Cheyenne")

ggplot() +
  geom_sf(data = KS_county) +
  geom_sf_text(
    data          = KS_less_Cheyenne,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue",
    nudge_x       = -0.1,
    nudge_y       = 0.1
  ) +
  geom_sf_text(
    data          = Cheyenne,
    aes(label     = NAME),
    size          = 2.5,
    color         = "red",
    nudge_y       = 0.2
  )

# With annotate()
ggplot() +
  geom_sf(data = KS_county) +
  geom_sf_text(
    data          = KS_less_Cheyenne,
    aes(label     = NAME),
    check_overlap = TRUE,
    size          = 3,
    color         = "blue",
    nudge_x       = -0.1,
    nudge_y       = 0.1
  ) +
  #--- use annotate to add texts on the map ---#
  annotate(
    geom  = "text",
    x     = -102,
    y     = 39.8,
    size  = 3,
    label = "Cheyennes",
    color = "red"
  )
```

### Arranging maps

```{r}
# Inset
three_counties <- KS_county %>% 
  filter(
    NAME %in% c("Cheyenne", "Sherman", "Wallace")
  )

g_three_counties <- three_counties %>% 
  ggplot() +
  geom_sf() +
  geom_sf_text(aes(label = NAME)) +
  theme_void()

g_three_counties
```

```{r}
# Convert the ggplot into a grob
grob_aoi <- ggplotGrob(g_three_counties)

# Check the class
class(grob_aoi)

# Create a map of Kansas
g_region <- ggplot() +
  geom_sf(data = KS_county) +
  geom_sf(
    data  = three_counties,
    fill  = "blue",
    color = "red",
    alpha = 0.5) +
  theme_void()

# Convert to a grob
grob_region <- ggplotGrob(g_region)

g_inset <- ggplot() +
  coord_equal(
    xlim   = c(0, 1),
    ylim   = c(0, 1),
    expand = FALSE
  )

g_inset
```

```{r}
g_inset +
  annotation_custom(
    grob_aoi,
    xmin = 0,
    xmax = 0.8,
    ymin = 0,
    ymax = 1
  )

g_inset +
  annotation_custom(
    grob_aoi,
    xmin = 0,
    xmax = 0.8,
    ymin = 0,
    ymax = 1
  ) +
  annotation_custom(
    grob_region,
    xmin = 0.6,
    xmax = 1,
    ymin = 0,
    ymax = 0.3
  )
```

```{r}
# Without border
g_inset +
  annotation_custom(
    grob_aoi,
    xmin = 0,
    xmax = 0.8,
    ymin = 0,
    ymax = 1
  ) +
  annotation_custom(
    grob_region,
    xmin = 0.6,
    xmax = 1,
    ymin = 0,
    ymax = 0.3
  ) +
  theme_void()
```

### Fine-tuning maps for publication

```{r}
gw_by_county <- st_join(
  KS_county,
  gw_KS_sf
) %>%
  as_tibble() %>%
  with_groups(
    c(COUNTYFP, year),
    summarize,
    af_used = sum(af_used, na.rm = TRUE)
  ) %>% 
  left_join(KS_county, ., by = "COUNTYFP") %>%
  filter(!is.na(year))

gw_by_county
```

```{r}
g_base <- gw_by_county %>% 
  ggplot() +
  geom_sf(aes(fill = af_used)) +
    facet_wrap(vars(year))

g_base
```

```{r}
# Setting the theme
g_base +
  theme_void()

theme_void()
```

```{r}
g_base +
  theme(axis.text = element_blank())
```

```{r}
theme_for_map <-
  theme(
    axis.ticks       = element_blank(),
    axis.text        = element_blank(),
    axis.line        = element_blank(),
    panel.border     = element_blank(),
    panel.grid       = element_line(
      color = "transparent"),
    panel.background = element_blank(),
    plot.background  = element_rect(
      fill  = "transparent",
      color = "transparent")
  )

g_base +
  theme_for_map
```

```{r}
theme_for_map_2 <- 
  theme_void() +
  theme(
    strip.background = element_rect(
      fill = "gray"
    )
  )

g_base +
  theme_for_map_2
```

```{r}
p <- gw_by_county %>%
  ggplot(aes(fill = af_used)) +
  geom_sf() +
  facet_grid(vars(year)) +
  theme_bw()

p

p +
  theme(
    panel.grid = element_line(
      color = "transparent"
    )
  )
```

```{r}
g_legend <- gw_by_county %>% 
  ggplot(aes(fill = af_used)) +
  geom_sf() +
  facet_wrap(vars(year)) +
  theme_void()

g_legend +
  scale_fill_viridis_c() +
  labs(
    fill = "Groundwater pumping (acre-feet)"
  ) +
  theme(
    legend.position   = "bottom",
    legend.key.height = unit(0.5, "cm"),
    legend.key.width  = unit(2, "cm"),
    legend.text       = element_text(
      size   = 12,
      family = "Times"
    ),
    legend.title      = element_text(
      size   = 12,
      family = "Times"
    )
  ) +
  guides(
    fill = guide_colorbar(title.position = "top")
  )
```

## 6.X OpenStreetMap with Dr. Dominic Royé

https://dominicroye.github.io/en/2018/accessing-openstreetmap-data-with-r/

```{r}
library(osmdata)
library(sf)
library(ggmap)

# The first five features
available_features() %>% 
  head()

# Amenities
available_tags("amenity") %>% 
  head()

# Shops
available_tags("shop") %>% 
  head()
```

```{r}
# Building the query
q <- getbb("Madrid") %>%
  opq() %>%
  add_osm_feature("amenity", "cinema")

# Query structure
q %>% 
str() 
```

```{r}
cinema <- osmdata_sf(q)
cinema
```

```{r}
# Our background map
mad_map <- get_map(
  getbb("Madrid"),
  maptype = "toner-background",
  source  = "stamen"
)

# Final map
mad_map %>%
  ggmap() +
  geom_sf(
    data        = cinema$osm_points,
    inherit.aes = FALSE,
    color       = "#238443",
    fill        = "#004529",
    alpha       = .5,
    size        = 4,
    shape       = 21
  ) +
  labs(
    x = "",
    y = ""
  )
```

```{r}
# Bounding box for the Iberian Peninsula
m <- c(-10, 30, 5, 46)

# Building the query
q <- m %>% 
  opq(timeout = 25*100) %>%
  add_osm_feature("name", "Mercadona") %>%
  add_osm_feature("shop", "supermarket")

# Query
mercadona <- osmdata_sf(q)

# Final map
ggplot(mercadona$osm_points) +
  geom_sf(
    color = "#08519c",
    fill  = "#08306b",
    alpha = .5,
    size  = 1,
    shape = 21
    ) +
  theme_void()
```
