---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (3rd ed.), Chapter 16 - Coordinate systems"
author: "Original Code: Hadley Wickham | Modifications: Antti Rask"
date: "2023-01-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 16 Coordinate systems

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 16.1 Linear coordinate systems

### 16.1.1 Zooming into a plot with coord_cartesian()

```{r}
base <- mpg %>% 
  ggplot(aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()

# Full dataset
base

# Scaling to 4--6 throws away data outside that range
base +
  scale_x_continuous(limits = c(4, 6))

# Zooming to 4--6 keeps all the data but only shows some of it
base +
  coord_cartesian(xlim = c(4, 6))
```

### 16.1.2 Flipping the axes with coord_flip()

```{r}
p <- mpg %>%
  ggplot(aes(displ, cty)) +
  geom_point() + 
  geom_smooth()

p

# Exchanging cty and displ rotates the plot 90 degrees, # but the smooth is fit to the rotated data.
mpg %>%
  ggplot(aes(cty, displ)) + 
  geom_point() + 
  geom_smooth()

# coord_flip() fits the smooth to the original data, and then rotates the output
p +
  coord_flip()
```

## 16.2 Non-linear coordinate systems

```{r}
rect <- tibble(x = 50, y = 50)
line <- tibble(x = c(1, 200), y = c(100, 1))

base <- ggplot(mapping = aes(x, y)) +
  geom_tile(
    data = rect,
    aes(width = 50, height = 50)
  ) + 
  geom_line(data = line) + 
  labs(
    x = NULL,
    y = NULL
  )

base +
  labs(title = "base")

base +
  coord_polar("x") +
  labs(title = 'coord_polar("x")')

base +
  coord_polar("y") +
  labs(title = 'coord_polar("y")')

base +
  coord_flip() +
  labs(title = "coord_flip")

base +
  coord_trans(y = "log10") +
  labs(title = 'coord_trans(y = "log10")')

base +
  coord_fixed() +
  labs(title = "coord_fixed()")
```

Munching:

1. We start with a line parameterised by its two endpoints:
```{r}
tbl <- tibble(
  r     = c(0, 1),
  theta = c(0, 3 / 2 * pi)
)

tbl %>%
  ggplot(aes(r, theta)) + 
  geom_line() + 
  geom_point(size = 2, color = "red")
```

2. We break it into multiple line segments, each with two endpoints:
```{r}
interp <- function(rng, n) {
  seq(rng[1], rng[2], length = n)
}

munched <- tibble(
  r     = interp(tbl$r, 15),
  theta = interp(tbl$theta, 15)
)

munched %>%
  ggplot(aes(r, theta)) + 
  geom_line() + 
  geom_point(size = 2, color = "red")
```

3. We transform the locations of each piece:
```{r}
transformed <- transform(
  munched,
  x = r * sin(theta),
  y = r * cos(theta)
)

transformed %>%
  ggplot(aes(x, y)) + 
  geom_path() + 
  geom_point(size = 2, color = "red") + 
  coord_fixed()
```

### 16.2.1 Transformations with coord_trans()

```{r, warning=FALSE}
library(scales)

# Linear model on original scale is poor fit
base <- diamonds %>% 
  ggplot(aes(carat, price)) + 
  stat_bin2d() + 
  geom_smooth(method = "lm") + 
  labs(
    x = NULL,
    y = NULL
  ) +
  theme(legend.position = "none")

base

# Better fit on log scale, but harder to interpret
p <- base +
  scale_x_log10() + 
  scale_y_log10()

p

# Fit on log scale, then backtransform to original.
# Highlights lack of expensive diamonds with large
# carats
pow10 <- exp_trans(10)

p +
  coord_trans(x = pow10, y = pow10)
```

### 16.2.2 Polar coordinates with coord_polar()

```{r}
base <- mtcars %>% 
  ggplot(aes(factor(1), fill = factor(cyl))) +
  geom_bar(width = 1) +
  theme(legend.position = "none") +
  scale_x_discrete(NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, expand = c(0, 0))

# Stacked barchart
base

# Pie chart
base +
  coord_polar(theta = "y")

# The bullseye chart
base +
  coord_polar()
```

### 16.2.3 Map Projections with coord_map()

```{r}
# Prepare a map of NZ
nzmap <- map_data("nz") %>%
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  labs(
    x = NULL,
    y = NULL
  )

# Plot it in cartesian coordinates
nzmap

# With the aspect ratio approximation
nzmap +
  # coord_quickmap -> coord_sf
  coord_sf()
```

```{r}
# Since coord_map() has been superseded by coord_sf, I've modified this code to use that instead
world   <- map_data("world")
finland <- map_data("world", "finland")

worldmap <- world %>% 
  ggplot(aes(long, lat, group = group))

worldmap +
  geom_path() +
  geom_path(data = finland, color = "blue") +
  # coord_map -> coord_sf
  coord_sf()

worldmap +
  geom_polygon(fill = "white", color = "black") +
  geom_polygon(data = finland, fill = "blue") +
  # coord_map -> coord_sf
  coord_sf()
```
