---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (3rd. ed.), Chapter 7 - Networks"
author: "Original Code: Hadley Wickham (except Thomas Lin Pedersen for ggraph and tidygraph) | Modifications: Antti Rask"
date: "2023-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 7 Networks

```{r}
library(tidyverse)
```

## 7.1 What is network data?

### 7.1.1 A tidy network manipulation API

```{r}
library(tidygraph)

graph <- play_erdos_renyi(n = 10, p = 0.2) %>%
  activate(nodes) %>%
  mutate(class = sample(letters[1:4], n(), replace = TRUE)) %>%
  activate(edges) %>%
  arrange(.N()$class[from])

graph
```

### 7.1.2 Conversion

```{r}
library(ggraph)

head(highschool)
```

```{r}
hs_graph <- as_tbl_graph(highschool, directed = FALSE)

hs_graph
```

```{r}
luv_clust <- hclust(dist(luv_colours[, 1:3]))

luv_graph <- as_tbl_graph(luv_clust)

luv_graph
```

### 7.1.3 Algorithms

```{r}
graph %>%
  activate(nodes) %>%
  mutate(centrality = centrality_pagerank()) %>%
  arrange(desc(centrality))
```

### 7.1.4 Want more?

https://tidygraph.data-imaginist.com/reference/index.html

## 7.2 Visualizing networks

### 7.2.1 Setting up the visualization

#### 7.2.1.1 Specifying a layout

```{r}
library(ggraph)

hs_graph %>%
  ggraph() + 
  geom_edge_link() + 
  geom_node_point()

hs_graph %>%
  ggraph(layout = "drl") + 
  geom_edge_link() + 
  geom_node_point()

hs_graph <- hs_graph %>% 
  activate(edges) %>% 
  mutate(edge_weights = runif(n()))

hs_graph %>%
  ggraph(layout = "stress", weights = edge_weights) + 
  geom_edge_link(aes(alpha = edge_weights)) + 
  geom_node_point() + 
  scale_edge_alpha_identity()
```

#### 7.2.1.2 Circularity

```{r}
luv_graph %>% 
  ggraph(layout = "dendrogram", circular = TRUE) + 
  geom_edge_link() + 
  coord_fixed()

luv_graph %>%
  ggraph(layout = "dendrogram") + 
  geom_edge_link() + 
  coord_polar() + 
  scale_y_reverse()
```

### 7.2.2 Drawing nodes

```{r}
hs_graph %>% 
  ggraph(layout = "stress") + 
  geom_edge_link() + 
  geom_node_point(
    aes(
      filter = centrality_degree() > 2, 
      color  = centrality_power()),
    size = 4
  )

luv_graph %>% 
  ggraph(layout = "treemap") + 
  geom_node_tile(aes(fill = depth))
```

### 7.2.3 Drawing edges

```{r}
graph %>% 
  ggraph(layout = "stress") +
  geom_edge_link(aes(alpha = after_stat(index)))

graph %>%
  ggraph(layout = "stress") + 
  geom_edge_link2(
    aes(color = node.class), 
    width = 3,
    lineend = "round"
  )

p <- hs_graph %>%
  ggraph(layout = "stress")

p +
  geom_edge_fan()

p + 
  geom_edge_parallel()

luv_graph %>%
  ggraph(, layout = "dendrogram", height = height) + 
  geom_edge_elbow()
```

#### 7.2.3.1 Clipping edges around the nodes

```{r}
p <- graph %>%
  ggraph(layout = "stress")

p +
  geom_edge_link(arrow = arrow()) +
  geom_node_point(aes(color = class), size = 8)

p +
  geom_edge_link(
    arrow = arrow(),
    start_cap = circle(5, "mm"),
    end_cap = circle(5, "mm")
  ) +
  geom_node_point(aes(color = class), size = 8)
```

#### 7.2.3.2 An edge is not always a line

```{r}
hs_graph %>% 
  ggraph(layout = "matrix", sort.by = node_rank_traveller()) +
  geom_edge_point()
```

### 7.2.4 Faceting

```{r}
p <- hs_graph %>% 
  ggraph(layout = "stress") + 
  geom_edge_link() + 
  geom_node_point()

p +
  facet_edges(vars(year))

p +
  facet_nodes(~ group_spinglass())
```

## 7.X ggraph and tidygraph

ggraph: https://ggraph.data-imaginist.com/index.html
tidygraph: https://tidygraph.data-imaginist.com/

### Layouts

#### The ggraph() and create_layout() functions

```{r}
library(ggraph)
library(tidygraph)

set_graph_style(plot_margin = margin(1, 1, 1, 1))
graph <- as_tbl_graph(highschool)

# Not specifying the layout - defaults to "auto"
graph %>% 
  ggraph() + 
  geom_edge_link(aes(color = factor(year))) + 
  geom_node_point()

graph %>%
  ggraph(layout = 'kk') + 
  geom_edge_link(aes(color = factor(year))) + 
  geom_node_point()

graph %>%
  ggraph(, layout = 'kk', maxiter = 100) + 
  geom_edge_link(aes(color = factor(year))) + 
  geom_node_point()
```

```{r}
layout <- create_layout(graph, layout = 'eigen')

layout %>% 
  ggraph() + 
  geom_edge_link(aes(color = factor(year))) + 
  geom_node_point()

head(layout)

attributes(layout)
```

#### Layouts abound

##### A note on circularity

```{r}
# An arc diagram
graph %>% 
  ggraph(layout = 'linear') + 
  geom_edge_arc(aes(color = factor(year)))

# A coord diagram
graph %>% 
  ggraph(layout = 'linear', circular = TRUE) + 
  geom_edge_arc(aes(color = factor(year))) + 
  coord_fixed()
```

```{r}
graph <- tbl_graph(flare$vertices, flare$edges)

# An icicle plot
graph %>%
  ggraph('partition') + 
  geom_node_tile(aes(fill = depth), size = 0.25)

# A sunburst plot
graph %>%
  ggraph('partition', circular = TRUE) + 
  geom_node_arc_bar(aes(fill = depth), size = 0.25) + 
  coord_fixed()
```

##### Node-edge diagram layouts

```{r}
graph <- as_tbl_graph(highschool) %>% 
  mutate(degree = centrality_degree())

map(c('stress', 'fr', 'lgl', 'graphopt'), function(layout) {
  ggraph(graph, layout = layout) + 
    geom_edge_link(aes(color = factor(year)), show.legend = FALSE) +
    geom_node_point() + 
    labs(caption = str_c('Layout: ', layout))
})
```

##### Hive plots

```{r}
graph <- graph %>% 
  mutate(
    friends = case_when(
      centrality_degree(mode = 'in') < 5 ~ 'few',
      TRUE ~ case_when(
        centrality_degree(mode = 'in') >= 15 ~ 'many',
        TRUE ~ 'medium'
      )
    )
  )

graph %>%
  ggraph('hive', axis = friends, sort.by = degree) + 
  geom_edge_hive(aes(color = factor(year))) + 
  geom_axis_hive(aes(color = friends), size = 2, label = FALSE) + 
  coord_fixed()
```

##### Focal layouts

```{r}
library(ggforce)

graph %>%
  ggraph( 'focus', focus = node_is_center()) + 
  geom_circle(aes(x0 = 0, y0 = 0, r = r), tibble(r = 1:5), colour = 'grey') + 
  geom_edge_link() + 
  geom_node_point() + 
  coord_fixed()
```

##### Hierarchical layouts

```{r}
graph <- tbl_graph(flare$vertices, flare$edges)

set.seed(1)

p <- graph %>% 
  ggraph('circlepack', weight = size)

p +
  geom_node_circle(aes(fill = depth), size = 0.25, n = 50) +
  coord_fixed()

p + 
  geom_edge_link() + 
  geom_node_point(aes(colour = depth)) +
  coord_fixed()

p <- graph %>%
  ggraph('treemap', weight = size)

p + 
  geom_node_tile(aes(fill = depth), size = 0.25)

p + 
  geom_edge_link() + 
  geom_node_point(aes(colour = depth))

ggraph(graph, 'tree') + 
  geom_edge_diagonal()

dendrogram <- hclust(dist(iris[, 1:4]))

dendrogram %>%
  ggraph('dendrogram', height = height) + 
  geom_edge_elbow()

dendrogram %>%
  ggraph('dendrogram', circular = TRUE) + 
  geom_edge_elbow() + 
  coord_fixed()

tree <- create_tree(100, 2, directed = FALSE) %>% 
  activate(edges) %>% 
  mutate(length = runif(n()))

tree %>%
  ggraph('unrooted', length = length) +
  geom_edge_link()
```

##### Matrix layouts

```{r}
graph <- create_notable('zachary')

graph %>%
  ggraph('matrix', sort.by = node_rank_leafsort()) + 
  geom_edge_point(mirror = TRUE) + 
  coord_fixed()

graph %>% 
  ggraph('matrix', sort.by = node_rank_spectral()) +
  geom_edge_point(mirror = TRUE) + 
  coord_fixed()
```

##### Fabric layouts

```{r}
graph %>%
  ggraph('fabric', sort.by = node_rank_fabric()) +
  geom_node_range(colour = 'grey') +
  geom_edge_span(end_shape = 'square') +
  coord_fixed()
```

### Nodes

```{r}
library(ggraph)
library(tidygraph)

set_graph_style(plot_margin = margin(1, 1, 1, 1))
gr <- as_tbl_graph(highschool)

gr %>%
  ggraph(layout = 'kk') +
  geom_point(aes(x = x, y = y))

head(create_layout(graph, layout = 'kk'))
```

#### geom_node_*()

```{r}
gr %>%
  ggraph(layout = 'kk') + 
  geom_node_point()

gr <- tbl_graph(flare$vertices, flare$edges)

gr %>%
  ggraph(, layout = 'partition') + 
  geom_node_tile(aes(y = -y, fill = depth))

gr %>%
  ggraph(layout = 'dendrogram', circular = TRUE) + 
  geom_edge_diagonal() + 
  geom_node_point(aes(filter = leaf)) + 
  coord_fixed()
```

#### The different node geoms

```{r}
graph <- create_notable('meredith') %>% 
  mutate(group = sample(c('A', 'B'), n(), TRUE))

graph %>%
  ggraph('stress') + 
  geom_node_voronoi(aes(fill = group), max.radius = 1) + 
  geom_node_point() + 
  geom_edge_link() + 
  coord_fixed()

gr %>%
  ggraph(layout = 'treemap', weight = size) + 
  geom_node_tile(aes(fill = depth))

l <- gr %>% 
  ggraph(layout = 'partition', circular = TRUE)

l +
  geom_node_arc_bar(aes(fill = depth)) + 
  coord_fixed()

l + geom_edge_diagonal() + 
  geom_node_point(aes(colour = depth)) + 
  coord_fixed()
```

### Edges

#### Meet the geom_edge_*() family

```{r}
library(ggraph)
library(rlang)
library(tidygraph)

set_graph_style(plot_margin = margin(1, 1, 1, 1))

hierarchy <- as_tbl_graph(hclust(dist(iris[, 1:4]))) %>% 
  
  mutate(Class = map_bfs_back_chr(node_is_root(), .f = function(node, path, ...) {
    
    if (leaf[node]) {
      
      as.character(iris$Species[as.integer(label[node])])
      
    } else {
      
      species <- unique(unlist(path$result))
      
      if (length(species) == 1) {
        
        species
        
      } else {
        
        NA_character_
        
      }
      
    }
    
  }))

hairball <- as_tbl_graph(highschool) %>% 
  mutate(
    year_pop = map_local(mode = 'in', .f = function(neighborhood, ...) {
      neighborhood %E>% pull(year) %>% table() %>% sort(decreasing = TRUE)
    }),
    pop_devel = map_chr(year_pop, function(pop) {
      if (length(pop) == 0 || length(unique(pop)) == 1) return('unchanged')
      switch(names(pop)[which.max(pop)],
             '1957' = 'decreased',
             '1958' = 'increased')
    }),
    popularity = map_dbl(year_pop, ~ .[1]) %|% 0
  ) %>%
  activate(edges) %>%
  mutate(year = as.character(year))
```

##### Link

```{r}
p <- hairball %>%
  ggraph(layout = 'stress') 

p + 
  geom_edge_link(aes(colour = year))
```

##### Fan

```{r}
p + 
  geom_edge_fan(aes(colour = year))
```

##### Parallel

```{r}
p +
  geom_edge_parallel(aes(colour = year))
```

##### Loops

```{r}
# let's make some of the student love themselves
loopy_hairball <- hairball %>% 
  bind_edges(tibble(from = 1:5, to = 1:5, year = rep('1957', 5)))

loopy_hairball %>%
ggraph(layout = 'stress') + 
  geom_edge_link(aes(colour = year), alpha = 0.25) + 
  geom_edge_loop(aes(colour = year))
```

##### Density

```{r}
p + 
  geom_edge_density(aes(fill = year)) + 
  geom_edge_link(alpha = 0.25)
```

##### Arcs

```{r}
hairball %>%
  ggraph(layout = 'linear') +
  geom_edge_arc(aes(colour = year))

hairball %>%
ggraph(layout = 'linear', circular = TRUE) + 
  geom_edge_arc(aes(colour = year)) + 
  coord_fixed()
```

##### Elbow

```{r}
p <- hierarchy %>%
  ggraph(layout = 'dendrogram', height = height)

p +
  geom_edge_elbow()
```

##### Diagonals

```{r}
p + 
  geom_edge_diagonal()
```

##### Bends

```{r}
p +
  geom_edge_bend()
```

##### Hive

```{r}
hairball %>%
ggraph(layout = 'hive', axis = pop_devel, sort.by = popularity) + 
  geom_edge_hive(aes(colour = year)) + 
  geom_axis_hive(label = FALSE) + 
  coord_fixed()
```

##### Span

```{r}
hairball %>%
  ggraph(layout = 'fabric', sort.by = node_rank_fabric()) + 
  geom_node_range(colour = 'grey') + 
  geom_edge_span(end_shape = 'circle') + 
  coord_fixed()
```

##### Point and tile

```{r}
p <- ggraph(hairball, layout = 'matrix', sort.by = bfs_rank())

p +
  geom_edge_point() + 
  coord_fixed()

p +
  geom_edge_tile() + 
  coord_fixed()
```

#### The three types of edge geoms

##### Base variant

```{r}
hairball %>%
  ggraph(layout = 'linear') + 
  geom_edge_arc(aes(colour = year, alpha = after_stat(index))) +
  scale_edge_alpha('Edge direction', guide = 'edge_direction')
```

##### 2-variant

```{r}
hierarchy %>%
ggraph(layout = 'dendrogram', height = height) +
  geom_edge_elbow2(aes(colour = node.Class))
```

#### Edge strength

```{r}
small_tree <- create_tree(5, 2)

ggraph(small_tree, 'dendrogram') + 
  geom_edge_elbow(strength = 0.75)

ggraph(small_tree, 'dendrogram') + 
  geom_edge_diagonal(strength = 0.5)
```

#### Decorating edges

```{r}
# Random names - I swear
simple <- create_notable('bull') %>% 
  mutate(name = c('Thomas', 'Bob', 'Hadley', 'Winston', 'Baptiste')) %>% 
  activate(edges) %>% 
  mutate(type = sample(c('friend', 'foe'), 5, TRUE))
```

##### Arrows

```{r}
simple %>% 
  ggraph(layout = 'graphopt') +
  geom_edge_link(arrow = arrow(length = unit(4, 'mm'))) +
  geom_node_point(size = 5)

simple %>%
  ggraph(layout = 'graphopt') + 
  geom_edge_link(
    arrow = arrow(length = unit(4, 'mm')), end_cap = circle(3, 'mm')
  ) +
  geom_node_point(size = 5)

simple %>%
  ggraph(layout = 'linear', circular = TRUE) +
  geom_edge_arc(
    arrow = arrow(length = unit(4, 'mm')), start_cap = circle(3, 'mm'), end_cap = circle(3, 'mm')
  ) +
  geom_node_point(size = 5) +
  coord_fixed()

simple %>%
  ggraph(layout = 'graphopt') +
  geom_edge_link(
    aes(start_cap = label_rect(node1.name), end_cap = label_rect(node2.name)),
    arrow = arrow(length = unit(4, 'mm'))
  ) + 
  geom_node_text(aes(label = name))
```

##### Labels

```{r}
simple %>%
  ggraph(layout = 'graphopt') +
  geom_edge_link(aes(label = type), arrow = arrow(length = unit(4, 'mm')), end_cap = circle(3, 'mm')) +
  geom_node_point(size = 5)

simple %>% 
  ggraph(layout = 'graphopt') +
  geom_edge_link(
    aes(label   = type),
    angle_calc  = 'along',
    label_dodge = unit(2.5, 'mm'),
    arrow       = arrow(length = unit(4, 'mm')),
    end_cap     = circle(3, 'mm')) + 
  geom_node_point(size = 5)
```

#### Connections

```{r}
flaregraph <- tbl_graph(flare$vertices, flare$edges)
from       <- match(flare$imports$from, flare$vertices$name)
to         <- match(flare$imports$to, flare$vertices$name)

flaregraph %>%
  ggraph(layout = 'dendrogram', circular = TRUE) +
  geom_conn_bundle(data = get_con(from = from, to = to), alpha = 0.1) +
  coord_fixed()
```


### Tidygraph interface

Original from: https://ggraph.data-imaginist.com/articles/tidygraph.html

#### NSE in layout specifications

```{r}
library(tidygraph)

graph <- as_tbl_graph(
  tibble(
    from = sample(5, 20, TRUE),
    to = sample(5, 20, TRUE),
    weight = runif(20)
  )
)

graph

```

```{r}
graph %>%
ggraph(layout = 'fr', weights = weight) +
  geom_edge_link() + 
  geom_node_point()

graph %>%
ggraph(layout = 'fr', weights = exp(weight)) +
  geom_edge_link() +
  geom_node_point()
```

#### Access to tidygraph algorithms in ggraph code

```{r}
graph <- create_notable('zachary')

graph %>%
  ggraph(layout = 'fr') + 
  geom_edge_link() + 
  geom_node_point(aes(size = centrality_pagerank())) + 
  theme(legend.position = 'bottom')

graph %>%
  ggraph('matrix', sort.by = node_rank_leafsort()) +
  geom_edge_point(aes(colour = centrality_edge_betweenness()), mirror = TRUE) +
  theme(legend.position = 'bottom')

graph %>%
  ggraph('fr') + 
  geom_edge_link() + 
  geom_node_point() + 
  facet_nodes(~ group_infomap())
```
