---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (3rd ed.), Chapter 11 - Colour scales and legends"
author: "Original Code: Hadley Wickham | Modifications: Antti Rask"
date: "2023-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 11 Colour scales and legends

```{r}
library(tidyverse)
library(conflicted)
  conflict_prefer("filter", "dplyr")
  conflict_prefer("select", "dplyr")
```

## 11.1 A little colour theory

### 11.1.1 Colour blindness

```{r}
library(colorBlindness)
library(viridis)

displayAllColors(rainbow(6))

displayAllColors(viridis(6))
```

## 11.2 Continuous colour scales

```{r}
library(munsell)

erupt <- faithfuld %>%
  ggplot(aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_x_continuous(NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, expand = c(0, 0)) +
  theme(legend.position = "none")

# This is extra (from the 2nd edition)

erupt

erupt +
  scale_fill_gradient(
    low = "white",
    high = "black"
  )

erupt +
  scale_fill_gradient(
    low  = mnsl("5G 9/2"),
    high = mnsl("5G 6/8")
  )
```

### 11.2.1 Particular palettes

```{r}
library(paletteer)
library(scico)
library(viridis)

erupt

erupt +
  scale_fill_viridis_c() +
  labs(title = "scale_fill_viridis_c()")

erupt +
  scale_fill_viridis_c(option = "magma") +
  labs(title = 'scale_fill_viridis_c(option = "magma")')

erupt +
  scale_fill_distiller() +
  labs(title = "scale_fill_distiller()")

erupt +
  scale_fill_distiller(palette = "RdPu") +
  labs(title = 'scale_fill_distiller(palette = "RdPu")')

erupt +
  scale_fill_distiller(palette = "YlOrBr") +
  labs(title = 'scale_fill_distiller(palette = "YlOrBr")')

# The default
erupt +
  scale_fill_scico(palette = "bilbao") +
  labs(title = 'scale_fill_scico(palette = "bilbao")')

erupt +
  scale_fill_scico(palette = "vik") +
  labs(title = 'scale_fill_scico(palette = "vik")')

erupt +
  scale_fill_scico(palette = "lajolla") +
  labs(title = 'scale_fill_scico(palette = "lajolla")')

erupt + 
  scale_fill_paletteer_c("viridis::plasma") +
  labs(title = 'scale_fill_paletteer_c("viridis::plasma")')

erupt +
  scale_fill_paletteer_c("scico::tokyo") +
  labs(title = 'scale_fill_paletteer_c("scico::tokyo")')
```

### 11.2.2 Robust recipes

```{r}
# These all produce the same plot

erupt

erupt +
  scale_fill_continuous()

erupt +
  scale_fill_gradient()
```

```{r}
erupt + 
  scale_fill_gradient(low = "grey", high = "brown") +
  labs(
    title = 'scale_fill_gradient(low = "grey", high = "brown")')

erupt + 
  scale_fill_gradient2(
    low      = "grey", 
    mid      = "white", 
    high     = "brown", 
    midpoint = .02
  ) +
  labs(
    title = 'scale_fill_gradient2(
    low = "grey", mid = "white",
    high = "brown", midpoint = .02
    )'
  )

erupt + 
  scale_fill_gradientn(colors = terrain.colors(7)) +
  labs(title = "scale_fill_gradientn(colors = terrain.colors(7))")
```

```{r}
library(munsell)

# Generate a ggplot with hue_slice()
hue_slice("5P") +  
  # Add arrows for annotation
  annotate( 
    geom  = "segment", 
    x     = c(7, 7), 
    y     = c(1, 10), 
    xend  = c(7, 7), 
    yend  = c(2, 9), 
    arrow = arrow(length = unit(2, "mm"))
  )

# Construct scale
erupt +
  scale_fill_gradient(
    low  = mnsl("5P 2/12"), 
    high = mnsl("5P 7/12")
)
```

```{r}
library(munsell)
library(colorspace)

erupt + 
  scale_fill_gradient2(
    low      = mnsl("5B 7/8"),
    high     = mnsl("5Y 7/8"),
    mid      = mnsl("N 7/0"),
    midpoint = .02
  ) +
  labs(title = 'scale_fill_gradient2(low = mnsl("5B 7/8"), high = mnsl("5Y 7/8"), mid = mnsl("N 7/0"),
    midpoint = .02)')

erupt +
  scale_fill_gradientn(colors = heat_hcl(7)) +
  labs(title = "scale_fill_gradientn(colors = heat_hcl(7))")

erupt +
  scale_fill_gradientn(colors = diverge_hcl(7)) +
  labs(title = "scale_fill_gradientn(colors = diverge_hcl(7))")

# Extra from the second edition
mid <- median(faithfuld$density)

erupt +
  scale_fill_gradient2(midpoint = mid) +
  labs(title = "scale_fill_gradient2(midpoint = mid)")
```

### 11.2.3 Missing values

```{r}
tbl <- tibble(x = 1, y = 1:5, z = c(1, 3, 2, NA, 5))

p <- tbl %>% 
  ggplot(aes(x, y, fill = z)) +
  geom_tile(linewidth = 5) +
  labs(
    x = NULL,
    y = NULL
  ) +
  scale_x_continuous(labels = NULL)

p

# Make missing colors invisible
p +
  scale_fill_gradient(na.value = NA)

p +
  scale_fill_gradient(na.value = "yellow")

# Extra from the second edition
p +
  scale_fill_gradient(
    low      = "black",
    high     = "white",
    na.value = "red"
  )
```

### 11.2.4 Limits, breaks, and labels

```{r}
library(scales)

toy <- tibble(
  const = 1,
  up    = 1:4,
  txt   = letters[1:4],
  big   = (1:4) * 1000,
  log   = c(2, 5, 10, 2000)
)

base <- toy %>% 
  ggplot(aes(up, up, fill = big)) + 
  geom_tile() + 
  labs(
    x = NULL,
    y = NULL
  ) 

base 

base +
  scale_fill_continuous(limits = c(0, 10000))

base +
  scale_fill_continuous(breaks = c(1000, 2000, 4000))

base +
  scale_fill_continuous(labels = label_dollar())
```

### 11.2.5 Legends

```{r}
base <- mpg %>% 
  ggplot(aes(cyl, displ, color = hwy)) +
  geom_point(size = 2)

base

# These next two give same results
base + 
  guides(color = guide_colorbar(reverse = TRUE)) +
  labs(title = "guides(color = guide_colorbar(reverse = TRUE))")

base + 
  scale_color_continuous(guide = guide_colorbar(reverse = TRUE)) +
  labs(title = "scale_color_continuous(guide = guide_colorbar(reverse = TRUE))")

base +
  guides(color = guide_colorbar(barheight = unit(2, "cm"))) +
  labs(title = 'guides(color = guide_colorbar(barheight = unit(2, "cm")))')

base + 
  guides(color = guide_colorbar(direction = "horizontal")) +
  labs(title = 'guides(color = guide_colorbar(direction = "horizontal")')
```

## 11.3 Discrete colour scales

```{r}
library(wesanderson)

tbl <- tibble(
  x = c("a", "b", "c", "d"),
  y = c(3, 4, 1, 2)
)

bars <- tbl %>%
  ggplot(aes(x, y, fill = x)) +
  geom_bar(stat = "identity") +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme(legend.position = "none")

# These three are identical
bars

bars + scale_fill_discrete()

bars + scale_fill_hue()
```

### 11.3.1 Brewer scales

```{r}
library(RColorBrewer)

display.brewer.all()
```

```{r}
bars +
    scale_fill_brewer(palette = "Set1") +
    labs(title = 'scale_fill_brewer(palette = "Set1")')

bars +
    scale_fill_brewer(palette = "Set2") +
    labs(title = 'scale_fill_brewer(palette = "Set2")')

bars +
    scale_fill_brewer(palette = "Accent") +
    labs(title = 'scale_fill_brewer(palette = "Accent")')
```

```{r}
# Bright colors work best with points
tbl <- data.frame(
  x = 1:3 + runif(30),
  y = runif(30),
  z = c("a", "b", "c")
) %>% as_tibble()

point <- tbl %>% 
  ggplot(aes(x, y, color = z)) +
  geom_point() +
  theme(legend.position = "none") +
  labs(
    x = NULL,
    y = NULL
  )

point +
  scale_color_brewer(palette = "Set1") +
  labs(title = 'scale_color_brewer(palette = "Set1")')

point +
  scale_color_brewer(palette = "Set2") +
  labs(title = 'scale_color_brewer(palette = "Set2")')

point +
  scale_color_brewer(palette = "Pastel1") +
  labs(title = 'scale_color_brewer(palette = "Pastel1")')
```

```{r}
# Subtler colors work better with areas
tbl <- tibble(x = 1:3, y = 3:1, z = c("a", "b", "c"))

area <- tbl %>% 
  ggplot(aes(x, y, fill = z)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  labs(
    x = NULL,
    y = NULL
  )

area +
  scale_fill_brewer(palette = "Set1") +
  labs(title = 'scale_fill_brewer(palette = "Set1")')

area +
  scale_fill_brewer(palette = "Set2") +
  labs(title = 'scale_fill_brewer(palette = "Set2")')

area +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = 'scale_fill_brewer(palette = "Pastel1")')
```

### 11.3.2 Hue and grey scales

```{r}
bars

bars +
    scale_fill_hue(c = 40) +
    labs(title = "scale_fill_hue(c = 40)")

bars +
    scale_fill_hue(h = c(180, 300)) +
    labs(title = "scale_fill_hue(h = c(180, 300))")

bars +
    scale_fill_grey() +
    labs(title = "scale_fill_grey()")

bars +
    scale_fill_grey(start = 0.5, end = 1) +
    labs(title = "scale_fill_grey(start = 0.5, end = 1)")

bars +
    scale_fill_grey(start = 0, end = 0.5) +
    labs(title = "scale_fill_grey(start = 0, end = 0.5)")
```

### 11.3.3 Paletteer scales

```{r}
library(paletteer)

bars +
  scale_fill_paletteer_d("rtist::vangogh") +
  labs(title = 'scale_fill_paletteer_d("rtist::vangogh")')

bars +
  scale_fill_paletteer_d("colorBlindness::paletteMartin") +
  labs(title = 'scale_fill_paletteer_d("colorBlindness::paletteMartin")')

bars +
  scale_fill_paletteer_d("wesanderson::FantasticFox1") +
  labs(title = 'scale_fill_paletteer_d("wesanderson::FantasticFox1")')
```

### 11.3.4 Manual scales

```{r}
bars + 
  scale_fill_manual(values = c("sienna1", "sienna4", "hotpink1", "hotpink4")) +
  labs(title = 'scale_fill_manual(values = c("sienna1", "sienna4", "hotpink1", "hotpink4"))')

bars + 
  scale_fill_manual(values = c("tomato1", "tomato2", "tomato3", "tomato4")) +
  labs(title = 'scale_fill_manual(values = c("tomato1", "tomato2", "tomato3", "tomato4"))')

bars + 
  scale_fill_manual(values = c("grey", "black", "grey", "grey")) +
  labs(title = 'scale_fill_manual(values = c("grey", "black", "grey", "grey"))')

# You can also use a named vector to specify colors to be assigned to each level which allows you to specify the levels in any order you like
bars + 
  scale_fill_manual(
    values = c(
      "d" = "grey",
      "c" = "grey",
      "b" = "black",
      "a" = "grey"
    )
  ) +
  labs(title = 'scale_fill_manual(values = c("d" = "grey", "c" = "grey", "b" = "black", "a" = "grey"))')

# Extra from the second edition
bars +
    scale_fill_manual(values = wes_palette("GrandBudapest1")) +
    labs(title = 'scale_fill_manual(values = wes_palette("GrandBudapest1"))')
bars +
    scale_fill_manual(values = wes_palette("Zissou1")) +
    labs(title = 'scale_fill_manual(values = wes_palette("Zissou1"))')

bars +
    scale_fill_manual(values = wes_palette("Rushmore1")) +
    labs(title = 'scale_fill_manual(values = wes_palette("Rushmore1"))')
```

### 11.3.5 Limits, breaks, and labels

```{r}
base_99 <- mpg %>%
  filter(year == 1999) %>%
  ggplot(aes(displ, hwy, color = fl)) +
  geom_point() 
  
base_08 <- mpg %>%
  filter(year == 2008) %>% 
  ggplot(aes(displ, hwy, color = fl)) +
  geom_point() 

base_99

base_08

base_99 + 
  lims(color = c("c", "d", "e", "p", "r")) +
  labs(title = 'lims(color = c("c", "d", "e", "p", "r"))')

base_08 +
  lims(color = c("c", "d", "e", "p", "r")) +
  labs(title = 'lims(color = c("c", "d", "e", "p", "r"))')

base_99 + 
  lims(
    x      = c(1, 7), 
    y      = c(10, 45), 
    color = c("c", "d", "e", "p", "r")
  ) +
  labs(title = 'lims(
    x      = c(1, 7), 
    y      = c(10, 45), 
    color = c("c", "d", "e", "p", "r")
  )')

base_08 + 
  lims(
    x      = c(1, 7), 
    y      = c(10, 45), 
    color = c("c", "d", "e", "p", "r")
  ) +
  labs(title = 'lims(
    x      = c(1, 7), 
    y      = c(10, 45), 
    color = c("c", "d", "e", "p", "r")
  )')

base_99 + 
  scale_color_discrete(
    limits = c("c", "d", "e", "p", "r"), 
    breaks = c("d", "p", "r"),
    labels = c("diesel", "premium", "regular")
  ) +
  labs(title = 'scale_color_discrete(
    limits = c("c", "d", "e", "p", "r"), 
    breaks = c("d", "p", "r"),
    labels = c("diesel", "premium", "regular")
  )')

base_99 + 
  lims(x = c(1, 7), y = c(10, 45)) +
  scale_color_discrete(
    limits = c("c", "d", "e", "p", "r"), 
    breaks = c("d", "p", "r"),
    labels = c("diesel", "premium", "regular")
  ) +
  labs(title = 'lims(x = c(1, 7), y = c(10, 45)) +
  scale_color_discrete(
    limits = c("c", "d", "e", "p", "r"), 
    breaks = c("d", "p", "r"),
    labels = c("diesel", "premium", "regular")
  )')

base_08 + 
  lims(x = c(1, 7), y = c(10, 45)) +
  scale_color_discrete(
    limits = c("c", "d", "e", "p", "r"), 
    labels = c("compressed", "diesel", "ethanol", "premium", "regular")
  ) +
  labs(title = 'lims(x = c(1, 7), y = c(10, 45)) +
  scale_color_discrete(
    limits = c("c", "d", "e", "p", "r"), 
    labels = c("compressed", "diesel", "ethanol", "premium", "regular")
  )')
```

### 11.3.6 Legends

```{r}
base <- mpg %>% 
  ggplot(aes(drv, fill = factor(cyl))) +
  geom_bar() 

base

base +
  guides(fill = guide_legend(ncol = 2))

base +
  guides(fill = guide_legend(ncol = 2, byrow = TRUE))

base +
  guides(fill = guide_legend(reverse = TRUE))
```

```{r}
base <- mpg %>% 
  ggplot(aes(displ, hwy, color = drv)) +
  geom_point(size = 4, alpha = .2, stroke = 0)

base +
  guides(color = guide_legend())

base +
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```

## 11.4 Binned colour scales

```{r}
library(viridis)

erupt +
  scale_fill_binned() +
  labs(title = "scale_fill_binned()")

erupt +
  scale_fill_steps() +
  labs(title = "scale_fill_steps()")

erupt +
  scale_fill_steps(n.breaks = 8) +
  labs(title = "scale_fill_steps(n.breaks = 8)")

erupt + 
  scale_fill_steps(low = "grey", high = "brown") +
  labs(title = 'scale_fill_steps(low = "grey", high = "brown")')

erupt + 
  scale_fill_steps2(
    low = "grey", 
    mid = "white", 
    high = "brown", 
    midpoint = .02
  ) +
  labs(title = 'scale_fill_steps2(
    low = "grey", 
    mid = "white", 
    high = "brown", 
    midpoint = .02
  )')

erupt +
  scale_fill_stepsn(n.breaks = 12, colors = terrain.colors(12)) +
  labs(title = "scale_fill_stepsn(n.breaks = 12, colors = terrain.colors(12))")

erupt + 
  scale_fill_stepsn(n.breaks = 9, colors = viridis(9)) +
  labs(title = "scale_fill_stepsn(n.breaks = 9, colors = viridis(9))")

erupt +
  scale_fill_stepsn(n.breaks = 9, colors = magma(9)) +
  labs(title = "scale_fill_stepsn(n.breaks = 9, colors = magma(9))")

erupt +
  scale_fill_stepsn(n.breaks = 9, colors = inferno(9)) +
  labs(title = "scale_fill_stepsn(n.breaks = 9, colors = inferno(9))")

erupt +
  scale_fill_fermenter(n.breaks = 9) +
  labs(title = "scale_fill_fermenter(n.breaks = 9)")

erupt +
  scale_fill_fermenter(n.breaks = 9, palette = "Oranges") +
  labs(title = 'scale_fill_fermenter(n.breaks = 9, palette = "Oranges")')

erupt +
  scale_fill_fermenter(n.breaks = 9, palette = "PuOr") +
  labs(title = 'scale_fill_fermenter(n.breaks = 9, palette = "PuOr")')
```

### 11.4.2 Legends

```{r}
base <- mpg %>% 
  ggplot(aes(cyl, displ, color = hwy)) +
  geom_point(size = 2) +
  scale_color_binned()

base

base +
  guides(color = guide_colorsteps(show.limits = TRUE)) +
  labs(title = "guides(color = guide_colorsteps(show.limits = TRUE))")
  
```

## 11.5 Date-time colour scales

```{r}
Sys.setlocale(category = "LC_ALL", locale = "us")

base <- economics %>% 
  ggplot(aes(psavert, uempmed, color = date)) + 
  geom_point() 

base

base + 
  scale_color_date(
    date_breaks = "142 months", 
    date_labels = "%b %Y"
  )
```

## 11.6 Alpha scales

```{r}
faithfuld %>% 
  ggplot(aes(waiting, eruptions, alpha = density)) +
  geom_raster(fill = "maroon") +
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0))
```

○## 11.7 Legend position

```{r}
base <- ggplot(toy, aes(up, up)) + 
  geom_point(aes(color = txt), size = 3) + 
  xlab(NULL) + 
  ylab(NULL)

base + 
  theme(legend.position = "left") +
  labs(title = 'theme(legend.position = "left")')

# The default
base + 
  theme(legend.position = "right") +
  labs(title = 'theme(legend.position = "right") (the default)')

base + 
  theme(legend.position = "bottom") +
  labs(title = 'theme(legend.position = "bottom")')

base + 
  theme(legend.position = "none") +
  labs(title = 'theme(legend.position = "none")')

base + 
  theme(
    legend.position = c(0, 1), 
    legend.justification = c(0, 1)
  ) +
  labs(title = 'theme(
    legend.position = c(0, 1), 
    legend.justification = c(0, 1)
  )')

base + 
  theme(
    legend.position = c(0.5, 0.5), 
    legend.justification = c(0.5, 0.5)
  ) +
  labs(title = 'theme(
    legend.position = c(0.5, 0.5), 
    legend.justification = c(0.5, 0.5)
  )')

base + 
  theme(
    legend.position = c(1, 0), 
    legend.justification = c(1, 0)
  ) +
  labs(title = 'theme(
    legend.position = c(1, 0), 
    legend.justification = c(1, 0)
  )')
```
