---
title: "Wickham, Hadley - ggplot2: Elegant Graphics for Data Analysis (3rd ed.), Chapter 22 - A case study"
author: "Antti Rask"
date: "2023-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 22 A case study

```{r}
library(tidyverse)
```

We’ll develop the extension in five phases (from the book):

1. We’ll start as simple as possible by using the existing geom_path(), pairing it with a new stat.

2. The new stat only allows fixed diameter and tension, so we’ll next allow these to be set as parameters.

3. A stat is a great place to start but has some fundamental restrictions so we’ll convert our work so far in to a proper geom.

4. Geoms can only use dimensions relative to the data, and can’t use absolute sizes like 2cm, so next we’ll show you how to draw the spring with grid.

5. We’ll finish up by providing a custom scale and legend to pair with the geom.

## 22.1 What is a spring?

```{r}
circle <- tibble(
  x     = sin(seq(0, 2 * pi, length.out = 100)),
  y     = cos(seq(0, 2 * pi, length.out = 100)),
  index = 1:100,
  type  = "circle"
)
spring      <- circle
spring$x    <- spring$x + seq(0, 1.5, length.out = 100)
spring$type <- "spring"

rbind(circle, spring) %>% 
  ggplot(aes(x, y, group = type, alpha = index)) +
  geom_path(show.legend = FALSE) +
  facet_wrap(vars(type), scales = "free_x")
```

## 22.2 Part 1: A stat

### 22.2.1 Building functionality

```{r}
library(rlang)

create_spring <- function(
  x,
  y,
  xend,
  yend,
  diameter = 1,
  tension  = 0.75,
  n        = 50
) {
  if (tension <= 0) {
    abort("`tension` must be larger than zero.")
  }
  if (diameter == 0) {
    abort("`diameter` can not be zero.")
  }
  if (n == 0) {
    abort("`n` must be greater than zero.")
  }
  # Calculate direct length of segment
  length <- sqrt((x - xend) ^ 2 + (y - yend) ^ 2)
  
  # Figure out how many revolutions and points we need
  n_revolutions <- length / (diameter * tension)
  n_points      <- n * n_revolutions
  
  # Calculate sequence of radians and x and y offset
  radians <- seq(0, n_revolutions * 2 * pi, length.out = n_points)
  x       <- seq(x, xend, length.out = n_points)
  y       <- seq(y, yend, length.out = n_points)
  
  # Create the new data
  tibble(
    x = cos(radians) * diameter/2 + x,
    y = sin(radians) * diameter/2 + y
  )
}

spring <- create_spring(
  x        = 4,
  y        = 2,
  xend     = 10,
  yend     = 6,
  diameter = 2,
  tension  = 0.75,
  n        = 50
)

spring %>%
  ggplot(aes(x, y)) +
  geom_path()
```

```{r}
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

StatSpring <- ggproto(
  "StatSpring",
  Stat,
  setup_data = function(data, params) {
    if (anyDuplicated(data$group)) {
      data$group <- paste(data$group, seq_len(nrow(data)), sep = "-")
    }
    data
  },
  compute_panel = function(
    data,
    scales,
    diameter = 1,
    tension  = 0.75,
    n        = 50
  ) {
    cols_to_keep <- setdiff(names(data), c("x", "y", "xend", "yend"))
    springs      <- lapply(seq_len(nrow(data)), function(i) {
      spring_path <- create_spring(
        data$x[i],
        data$y[i], 
        data$xend[i],
        data$yend[i], 
        diameter = diameter, 
        tension  = tension, 
        n        = n
      )
      cbind(spring_path, unclass(data[i, cols_to_keep]))
    })
    do.call(rbind, springs)
  },
  required_aes = c("x", "y", "xend", "yend")
)
```

### 22.2.2 Methods

```{r}
Stat
```

```{r}
StatSpring$compute_panel
```

```{r}
StatSpring$setup_data
```

```{r}
StatSpring$required_aes
```

### 22.2.3 Constructors

```{r}
geom_spring <- function(
  mapping     = NULL,
  data        = NULL, 
  stat        = "spring",
  position    = "identity", 
  ..., 
  diameter    = 1, 
  tension     = 0.75,
  n           = 50, 
  arrow       = NULL, 
  lineend     = "butt", 
  linejoin    = "round",
  na.rm       = FALSE, 
  show.legend = NA, 
  inherit.aes = TRUE
) {
  layer(
    data        = data,
    mapping     = mapping,
    stat        = stat,
    geom        = GeomPath,
    position    = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params      = list(
      diameter = diameter,
      tension  = tension,
      n        = n,
      arrow    = arrow,
      lineend  = lineend,
      linejoin = linejoin,
      na.rm    = na.rm,
      ...
    )
  )
}

some_data <- tibble(
  x     = runif(5, max = 10),
  y     = runif(5, max = 10),
  xend  = runif(5, max = 10),
  yend  = runif(5, max = 10),
  class = sample(letters[1:2], 5, replace = TRUE)
)

some_data %>%
  ggplot(aes(x, y, xend = xend, yend = yend)) +
  geom_spring()

some_data %>%
  ggplot(aes(x, y, xend = xend, yend = yend, color = class)) +
  geom_spring(linewidth = 1) +
  facet_wrap(vars(class))
```

```{r}
stat_spring <- function(
  mapping     = NULL,
  data        = NULL,
  geom        = "path", 
  position    = "identity",
  ...,
  diameter    = 1,
  tension     = 0.75, 
  n           = 50,
  na.rm       = FALSE,
  show.legend = NA, 
  inherit.aes = TRUE
) {
  layer(
    data        = data, 
    mapping     = mapping, 
    stat        = StatSpring, 
    geom        = geom, 
    position    = position, 
    show.legend = show.legend, 
    inherit.aes = inherit.aes, 
    params      = list(
      diameter = diameter, 
      tension  = tension, 
      n        = n, 
      na.rm    = na.rm, 
      ...
    )
  )
}

some_data %>%
  ggplot(aes(x, y, xend = xend, yend = yend, color = class)) + 
  stat_spring(geom = "point", n = 15) +
  facet_wrap(vars(class))
```

### 22.2.4 Post-mortem

```{r}
ggplot() + 
  geom_spring(aes(x = 0, y = 0, xend = 3, yend = 20))

ggplot() + 
  geom_spring(aes(x = 0, y = 0, xend = 100, yend = 80))
```

## 22.3 Part 2: Adding aesthetics

```{r}
StatSpring <- ggproto(
  "StatSpring",
  Stat, 
  setup_data = function(data, params) {
    if (anyDuplicated(data$group)) {
      data$group <- paste(data$group, seq_len(nrow(data)), sep = "-")
    }
    data
  },
  compute_panel = function(data, scales, n = 50) {
    cols_to_keep  <- setdiff(names(data), c("x", "y", "xend", "yend"))
    springs       <- lapply(seq_len(nrow(data)), function(i) {
      spring_path <- create_spring(
        data$x[i],
        data$y[i],
        data$xend[i],
        data$yend[i],
        data$diameter[i],
        data$tension[i],
        n
      )
      cbind(spring_path, unclass(data[i, cols_to_keep]))
    })
    do.call(rbind, springs)
  },
  required_aes = c("x", "y", "xend", "yend"),
  optional_aes = c("diameter", "tension")
)

geom_spring <- function(
  mapping     = NULL,
  data        = NULL,
  stat        = "spring", 
  position    = "identity",
  ...,
  n           = 50,
  arrow       = NULL,
  lineend     = "butt",
  linejoin    = "round",
  na.rm       = FALSE,
  show.legend = NA,
  inherit.aes = TRUE
) {
  layer(
    data        = data, 
    mapping     = mapping, 
    stat        = stat, 
    geom        = GeomPath, 
    position    = position, 
    show.legend = show.legend, 
    inherit.aes = inherit.aes, 
    params      = list(
      n        = n, 
      arrow    = arrow, 
      lineend  = lineend, 
      linejoin = linejoin, 
      na.rm    = na.rm, 
      ...
    )
  )
}

some_data <- tibble(
  x        = runif(5, max = 10),
  y        = runif(5, max = 10),
  xend     = runif(5, max = 10),
  yend     = runif(5, max = 10),
  class    = sample(letters[1:2], 5, replace = TRUE),
  tension  = runif(5),
  diameter = runif(5, 0.5, 1.5)
)

p <- some_data %>%
  ggplot(aes(x, y, xend = xend, yend = yend))

p +
  geom_spring(aes(tension = tension, diameter = diameter))

p +
  geom_spring(diameter = 0.5)
```

## 22.4 Part 3: A geom

### 22.4.1 Geom extensions

```{r}
# Using the ... in this context caused this error: Error: '...' used in an incorrect context. I'm removing them from this bit of code, but realize that it might cause other problems down the line.

GeomSpring <- ggproto(
  "GeomSpring",
  GeomPath,
  # ...,
  setup_data = function(data, params) {
    cols_to_keep  <- setdiff(names(data), c("x", "y", "xend", "yend"))
    springs       <- lapply(seq_len(nrow(data)), function(i) {
      spring_path <- create_spring(
        data$x[i],
        data$y[i],
        data$xend[i],
        data$yend[i], 
        diameter = data$diameter[i],
        tension  = data$tension[i], 
        n        = params$n
      )
      spring_path       <- cbind(spring_path, unclass(data[i, cols_to_keep]))
      spring_path$group <- i
      spring_path
    })
    do.call(rbind, springs)
  },
  # ...
)

GeomSpring <- ggproto(
  "GeomSpring",
  Geom,
  setup_data = function(data, params) {
    if (is.null(data$group)) {
      data$group <- seq_len(nrow(data))
    }
    if (anyDuplicated(data$group)) {
      data$group <- paste(data$group, seq_len(nrow(data)), sep = "-")
    }
    data
  },
  draw_panel = function(
    data,
    panel_params,
    coord,
    n         = 50,
    arrow     = NULL,
    lineend   = "butt",
    linejoin  = "round",
    linemitre = 10,
    na.rm     = FALSE
  ) {
    cols_to_keep  <- setdiff(names(data), c("x", "y", "xend", "yend"))
    springs       <- lapply(seq_len(nrow(data)), function(i) {
      spring_path <- create_spring(
        data$x[i],
        data$y[i],
        data$xend[i],
        data$yend[i],
        data$diameter[i],
        data$tension[i],
        n
      )
      cbind(spring_path, unclass(data[i, cols_to_keep]))
    })
    springs <- do.call(rbind, springs)
    GeomPath$draw_panel(
      data         = springs, 
      panel_params = panel_params, 
      coord        = coord, 
      arrow        = arrow, 
      lineend      = lineend, 
      linejoin     = linejoin, 
      linemitre    = linemitre, 
      na.rm        = na.rm
    )
  },
  required_aes = c("x", "y", "xend", "yend"),
  default_aes  = aes(
    colour     = "black", 
    linewidth  = 0.5, 
    linetype   = 1L, 
    alpha      = NA, 
    diameter   = 1, 
    tension    = 0.75
  )
)

geom_spring <- function(
  mapping     = NULL,
  data        = NULL,
  stat        = "identity", 
  position    = "identity",
  ...,
  n           = 50,
  arrow       = NULL, 
  lineend     = "butt",
  linejoin    = "round",
  na.rm       = FALSE,
  show.legend = NA,
  inherit.aes = TRUE
) {
  layer(
    data        = data, 
    mapping     = mapping, 
    stat        = stat, 
    geom        = GeomSpring, 
    position    = position, 
    show.legend = show.legend, 
    inherit.aes = inherit.aes, 
    params      = list(
      n        = n, 
      arrow    = arrow, 
      lineend  = lineend, 
      linejoin = linejoin, 
      na.rm    = na.rm, 
      ...
    )
  )
}

some_data %>%
  ggplot(aes(x, y, xend = xend, yend = yend)) +
  geom_spring(diameter = 0.5)
```

## 22.5 Part 4: A grid grob

### 22.5.1 Grid in 5 minutes

#### 22.5.1.4 Example

```{r}
library(grid)

surpriseGrob <- function(
  x,
  y,
  size, 
  default.units = "npc", 
  name          = NULL,
  gp            = gpar(), 
  vp            = NULL
) {
  
  # Check if input needs to be converted to units
  if (!is.unit(x)) {
    x <- unit(x, default.units)
  }
  if (!is.unit(y)) {
    y <- unit(y, default.units)
  }
  if (!is.unit(size)) {
    size <- unit(size, default.units)
  }
  # Construct our surprise grob subclass as a gTree
  gTree(
    x    = x, 
    y    = y, 
    size = size, 
    name = name, 
    gp   = gp, 
    vp   = vp, 
    cl   = "surprise"
  )
}
makeContent.surprise <- function(x) {
  x_pos <- x$x
  y_pos <- x$y
  size  <- convertWidth(x$size, unitTo = "cm", valueOnly = TRUE)
  
  # Figure out if the given sizes are bigger or smaller than 5 cm
  circles <- size < 5
  
  # Create a circle grob for the small ones
  if (any(circles)) {
    circle_grob <- circleGrob(
      x = x_pos[circles], 
      y = y_pos[circles], 
      r = unit(size[circles] / 2, "cm")
    )
  } else {
    circle_grob <- nullGrob()
  }
  
  # Create a rect grob for the large ones
  if (any(!circles)) {
    square_grob <- rectGrob(
      x      = x_pos[!circles], 
      y      = y_pos[!circles], 
      width  = unit(size[!circles], "cm"),
      height = unit(size[!circles], "cm")
    )
  } else {
    square_grob <- nullGrob()
  }
  
  # Add the circle and rect grob as childrens of our input grob
  setChildren(x, gList(circle_grob, square_grob))
}

# Create an instance of our surprise grob defining to object with different
# sizes
gr <- surpriseGrob(
  x    = c(0.25, 0.75),
  y    = c(0.5, 0.5),
  size = c(0.1, 0.4)
)

# Draw it
grid.newpage()
grid.draw(gr)
```

### 22.5.2 The springGrob

```{r}
springGrob <- function(
  x0            = unit(0, "npc"),
  y0            = unit(0, "npc"), 
  x1            = unit(1, "npc"),
  y1            = unit(1, "npc"), 
  diameter      = unit(0.1, "npc"),
  tension       = 0.75,
  n             = 50,
  default.units = "npc",
  name          = NULL, 
  gp            = gpar(),
  vp            = NULL
) {
  if (!is.unit(x0)) x0 <- unit(x0, default.units)
  if (!is.unit(x1)) x1 <- unit(x1, default.units)
  if (!is.unit(y0)) y0 <- unit(y0, default.units)
  if (!is.unit(y1)) y1 <- unit(y1, default.units)
  if (!is.unit(diameter)) diameter <- unit(diameter, default.units)
  gTree(
    x0       = x0,
    y0       = y0,
    x1       = x1,
    y1       = y1,
    diameter = diameter,
    tension  = tension,
    n        = n,
    name     = name,
    gp       = gp,
    vp       = vp, 
    cl       = "spring"
  )
}

makeContent.spring <- function(x) {
  x0       <- convertX(x$x0, "mm", valueOnly = TRUE)
  x1       <- convertX(x$x1, "mm", valueOnly = TRUE)
  y0       <- convertY(x$y0, "mm", valueOnly = TRUE)
  y1       <- convertY(x$y1, "mm", valueOnly = TRUE)
  diameter <- convertWidth(x$diameter, "mm", valueOnly = TRUE)
  tension  <- x$tension
  n        <- x$n
  springs  <- lapply(seq_along(x0), function(i) {
    cbind(
      create_spring(
        x0[i],
        y0[i],
        x1[i],
        y1[i],
        diameter[i],
        tension[i],
        n
      ),
      id = i
    )
  })
  springs      <- do.call(rbind, springs)
  spring_paths <- polylineGrob(
    springs$x,
    springs$y,
    springs$id,
    default.units = "mm",
    gp = x$gp
  )
  setChildren(x, gList(spring_paths))
}

springs <- springGrob(
  x0       = c(0, 0),
  y0       = c(0, 0.5),
  x1       = c(1, 1),
  y1       = c(1, 0.5),
  diameter = unit(c(1, 3), "cm"),
  tension  = c(0.2, 0.7)
)

grid.newpage()
grid.draw(springs)
```

### 22.5.3 The last GeomSpring

```{r}
library(rlang)

GeomSpring <- ggproto(
  "GeomSpring",
  Geom,
  setup_params = function(data, params) {
    if (is.null(params$n)) {
      params$n <- 50
    } else if (params$n <= 0) {
      abort("Springs must be defined with `n` greater than 0")
    }
    params
  },
  draw_panel = function(
    data,
    panel_params,
    coord,
    n       = 50,
    lineend = "butt",
    na.rm   = FALSE
  ) {
    data <- remove_missing(data, na.rm = na.rm,
                           c("x", "y", "xend", "yend", "linetype", "size"),
                           name = "geom_spring")
    if (is.null(data) || nrow(data) == 0) return(zeroGrob())
    if (!coord$is_linear()) {
      warn("spring geom only works correctly on linear coordinate systems")
    }
    coord <- coord$transform(data, panel_params)
    return(
      springGrob(
        coord$x,
        coord$y,
        coord$xend,
        coord$yend,
        default.units = "native",
        diameter      = unit(coord$diameter, "cm"),
        tension       = coord$tension,
        n             = n,
        gp            = gpar(
          col     = alpha(coord$colour, coord$alpha),
          lwd     = coord$size * .pt,
          lty     = coord$linetype,
          lineend = lineend
        )
      )
    )
  },
  required_aes = c("x", "y", "xend", "yend"),
  default_aes = aes(
    colour    = "black", 
    size      = 0.5, 
    linetype  = 1L, 
    alpha     = NA, 
    diameter  = 0.35, 
    tension   = 0.75
  )
)
geom_spring <- function(
  mapping     = NULL,
  data        = NULL,
  stat        = "identity",
  position    = "identity",
  ...,
  n           = 50,
  lineend     = "butt", 
  na.rm       = FALSE,
  show.legend = NA,
  inherit.aes = TRUE
) {
  layer(
    data        = data, 
    mapping     = mapping, 
    stat        = stat, 
    geom        = GeomSpring, 
    position    = position, 
    show.legend = show.legend, 
    inherit.aes = inherit.aes, 
    params      = list(
      n       = n, 
      lineend = lineend, 
      na.rm   = na.rm, 
      ...
    )
  )
}

some_data %>%
  ggplot() + 
  geom_spring(
    aes(x, y, xend = xend, yend = yend, diameter = diameter, tension = tension)
  )
```

## 22.6 Part 5: Scales

### 22.6.1 Scaling

```{r}
library(grid)
library(rlang)

scale_tension_continuous <- function(..., range = c(0.1, 1)) {
  continuous_scale(
    aesthetics = "tension", 
    scale_name = "tension_c", 
    palette = scales::rescale_pal(range), 
    ...
  )
}

scale_tension <- scale_tension_continuous

scale_tension_discrete <- function(...) {
  abort("Tension cannot be used with discrete data")
}

scale_diameter_continuous <- function(..., range = c(0.25, 0.7), unit = "cm") {
  range <- convertWidth(unit(range, unit), "cm", valueOnly = TRUE)
  continuous_scale(
    aesthetics = "diameter", 
    scale_name = "diameter_c", 
    palette    = scales::rescale_pal(range), 
    ...
  )
}
scale_diameter <- scale_diameter_continuous
scale_tension_discrete <- function(...) {
  abort("Diameter cannot be used with discrete data")
}

some_data %>%
  ggplot(aes(x, y, xend = xend, yend = yend, tension = tension, diameter = diameter)) +
  geom_spring() + 
  scale_tension(range = c(0.1, 5))
```

### 22.6.2 draw_key_spring

```{r}
draw_key_point
```

```{r}
draw_key_spring <- function(data, params, size) {
  springGrob(
    x0       = 0,
    y0       = 0,
    x1       = 1,
    y1       = 1,
    diameter = unit(data$diameter, "cm"), 
    tension  = data$tension, gp = gpar(
      col = alpha(data$colour %||% "black", data$alpha),
      lwd = (data$size %||% 0.5) * .pt,
      lty = data$linetype %||% 1
    ),
    vp = viewport(clip = "on")
  )
}

GeomSpring$draw_key <- draw_key_spring

p <- some_data %>%
  ggplot(
    aes(
      x,
      y,
      xend     = xend,
      yend     = yend,
      tension  = tension,
      diameter = diameter
    )
  ) +
  geom_spring() +
  scale_tension(range = c(0.1, 5))

p

p + 
  theme(legend.key.size = unit(1, "cm"))

some_data %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend, colour = class)) + 
  geom_spring() + 
  theme(legend.key.size = unit(1, "cm"))
```
